Free volume theory explains the unusual behavior of viscosity in a non-confluent tissue during morphogenesis

Version 1

Abstract

A recent experiment on zebrafish blastoderm morphogenesis showed that the viscosity (η) of a non-confluent embryonic tissue grows sharply until a critical cell density (ϕS). The increase in η up to ϕS is similar to the behavior observed in several glass forming materials, which suggests that the cell dynamics is sluggish or glass-like. Surprisingly, η is a constant above ϕS. To determine the mechanism of this unusual dependence of η on ϕ, we performed extensive simulations using an agent-based model of a dense non-confluent two-dimensional tissue. We show that polydispersity in the cell size, and the propensity of the cells to deform, results in the saturation of the available free area per cell beyond a critical density. Saturation in the free space not only explains the viscosity plateau above ϕS but also provides a relationship between equilibrium geometrical packing to the dramatic increase in the relaxation dynamics.

Introduction

There is great interest in characterizing the mechanical and dynamical properties of embryonic tissues because they regulate embryo development [1–5]. Measurements of bulk properties, such as viscosity and elastic modulus, and the dynamics of individual cells through imaging techniques, have been interpreted by adapting concepts developed to describe phase transitions, glass transition, and active matter [6–9]. Several experiments have shown that during embryo morphogenesis, material properties of the tissues change dramatically [10–14]. Of relevance to our study is a remarkable finding that provided evidence that a phase transition (PT) occurs during zebrafish blastoderm morphogenesis, which was analyzed using rigidity percolation theory [14–17]. The authors also estimated the viscosity (η) of the blastoderm tissue using the micropipette aspiration technique [13, 18]. It was found that change in η is correlated with cell connectivity (⟨C⟩), rising sharply over a narrow range of ⟨C⟩. Surprisingly, a single geometrical quantity, the cell-cell contact topology controls both the rigidity PT as well as changes in η in this non-confluent tissue, thus linking equilibrium and transport properties.

Here, we focus on two pertinent questions that arise from the experiments on zebrafish blastoderm. First, experiments [14] showed that η increases as a function of the packing fraction (ϕ) till ϕ ≤ 0.87. The dependence of η on ϕ follows the well-known Vogel-Fulcher-Tamman (VFT) law [19], which predicts that η grows monotonically with ϕ. The VFT law, which is commonly used to analyze the viscosity of a class of glass forming materials [20], is given by η ~ exp where ϕ0 is a constant.Surprisingly, for packing fraction, ϕ ≥ ϕS ≈ 0.90, η deviates from the VFT law and is independent of ϕ, which cannot be explained using conventional theories for glasses [8, 21]. Second, the experimental data [14] was interpreted using equilibrium rigidity percolation theory [15–17] for an embryonic tissue in which cells undergo random cell divisions. A priori it is unclear why equilibrium concepts should hold in zebrafish morphogenesis, which one would expect is controlled by non-equilibrium processes such as self-propulsion, growth and cell division.

We show that the two conundrums (saturation of η at high densities and the use of equilibrium statistical mechanics in a growing system to explain phase transition) may be rationalized by (i) assuming that the interactions between the cells are soft, (ii) the cell sizes are highly heterogeneous (polydisperse), which is the case in zebrafish blastoderm. Using an agent-based (particle) simulation model of a two-dimensional non-confluent tissue, we explore the consequences of varying ϕ of interacting self-propelled polydisperse soft cells, on η. The central results of our study are: (i) The calculated effective viscosity (for details see Appendix F), for the polydisperse cell system, shows that for ϕ ≤ ϕS ≈ 0.90 the increase in viscosity follows the VFT law. Just as in experiments, η is essentially independent of ϕ at high (≥ ϕS) densities. (ii) The unusual dependence of η on ϕ is quantitatively explained using available free area (ϕfree). A given cell requires free area to move. The free area is created by movement of the neighboring cells. One would intuitively expect that the ϕfree should decrease with increasing cell density, due to cell jamming, which should and should slow down the overall dynamics. Indeed, we find that ϕfree decreases with increasing packing fraction (ϕ) until ϕS. When ϕ exceeds ϕS, the free area ϕfree saturates because the soft cells interpenetrate each other, resulting in the collective dynamics of cells becoming independent of ϕ for ϕ ≥ ϕS. As a consequence η saturates at high ϕ. (iii) Cells whose sizes are comparable to the available free area move almost like a particle in a liquid. The motility of small sized cells facilitates adjacent cells to move through multi-cell rearrangement even in a highly jammed environment. The facilitation mechanism, invoked in glassy systems [22] allows large cells to move with low mobility. A cascade of such facilitation processes enable all the cells to remain dynamic even above the onset density of the phase transition. (iv) We find that η (or relaxation time) does not depend on the waiting time for measurements even in the regime where viscosity saturates. In other words, there is no evidence of aging even in the regime where viscosity saturates. The tissue remains ergodic [23] at all densities. The cell-based simulations, which reproduce the salient experimental features, may be described using equilibrium statistical mechanics, thus providing credence to the use of cell contact mechanics to describe both rigidity PT and dynamics in an evolving non-confluent tissue [14].

Results

Experimental results

We first describe the experimental observations, which serve as the basis for carrying out the agent-based simulations. Fig. 1(A) shows the bright-field images of distinct stages during zebrafish morphogenesis. A 2D section of zebrafish blastoderm (Fig. 1 (B)) shows that there is considerable dispersion in cell sizes. The statistical properties of the cell sizes are shown in Fig. 6 (D). Fig. 1 (C) shows that η increases sharply over a narrow density range, and saturates when ϕ exceeds ϕS ≈ 0.90.

Structure and viscosity of non-confluent tissues:

(A) Bright-field single-plane images of an exemplary embryo of zebrafish before (t = −60 min), at the onset (t = 0 min), and after blastoderm spreading (t = 60 min). (B) Snapshot of 2D confocal sections at the 1st–2nd deep-cell layer of the blastoderm at t = 60 min. (A) and (B) are taken from [14]. (C) Viscosity η of zebrafish blastoderm as a function of ϕ in a log-linear scale using the data from [14]. The dashed line is the fit to VFT equation. Note that η does not change significantly beyond ϕ ≥ 0.87. (D) A typical snapshot taken from cell-based simulations for ϕ = 0.93. Cells are colored according to their radii (in μm) (color bar shown on the right). (E) The pair correlation function, g(r), as a function of r for ϕ = 0.93. The vertical dashed line is the position of the first peak (rmax = 17.0 μm). The pair correlation function does not exhibit signs of long-range order. Scale bars in (A) is 100 μm and in (B) is 50 μm.

Saturation in viscosity and relaxation time:

(A) Effective viscosity as a function of ϕ, with the dashed line being the fit to VFT equation. The inset shows at high ϕ. The dashed line in the inset is the expected behavior assuming that the VFT relation holds at all ϕ. (B) The self-intermediate scattering function Fs(q, t) as a function of t for 0.70 ≤ ϕ ≤ 0.905. The dashed line corresponds to . (C) A similar plot for ϕ > 0.905. (D) The logarithm of the relaxation time τα(s) as a function of ϕ. The VFT fit is given by the dashed line. The inset shows a zoomed-in view for ϕ ≥ ϕS.

Spectrum of relaxation times:

(A) Scatter plot of relaxation times τα(s) as a function of cell radius. From top to bottom, the plot corresponds to decreasing ϕ. The vertical dashed line is for Ri = 4.25 μm, beyond which the τα changes sharply at high packing fractions. (B) Histogram P (ln(τα)) as a function of ln(τα). Beyond ϕ = 0.90 (ϕS), the histogram peaks do not shift substantially towards a high τα values. (C) For ϕ ≤ ϕS P (ln(τα)) (scaled by P max(ln(τα))) falls on a master curve, as described in the main text. (D) Same as (C) except the results are for ϕ > 0.90. The data deviates from the Gaussian fit, shown by the dashed line.

Density dependent cell-cell overlap :

(A) Probability of overlap (hij) between two cells, P (hij), for various ϕ values. The peak in the distribution function shifts to higher values as ϕ increases. (B) Mean ⟨hij⟩ = ∫dhijP (hij) as a function of ϕ. Inset shows a pictorial illustration of h12 between two cells with radii R1 and R2 at a distance r12.

Changes in free area fraction with

ϕ: (A) Voronoi tessellation of cells for ϕ = 0.93 for a single realization. The orange circles represent actual cell sizes. The blue polygons show the Voronoi cell size. (B) Distribution of Voronoi cell size A as a function of ϕ. (C) Mean Voronoi cell size ⟨A⟩ as a function of ϕ. A zoomed-in view for ϕ > 0.860 is shown in the inset. (D) Distribution of free area P (Afree) for all ϕ. The vertical blue dashed line shows that the maximum in the distribution is at Afree ~ 50μm2. (E) Free area fraction ϕfree as a function of ϕ. Note that ϕfree saturates beyond ϕ = 0.90. An expanded view of the saturated region is shown in the right panel of (E).

Area distribution of the cells:

(A) Simulation snapshot for monodisperse cell system. The number of cells in the two-dimensional periodic box is N = 500. (B) Pair correlation function, g(r), as a function of r. There is clear evidence of order, as reflected in the sharp peaks at regular intervals, which reflects the packing in (A). (C) A schematic picture of polydisperse cell system from the simulations. Color bar on the right shows the scale of radii in μm. There is no discernible order. (D) Distribution of cell area extracted from experiment during morphogenesis of Zebrafish blastoderm (extracted from Fig. S2 (A)) [14]. (E) Same as (D) except, P (Ai), used in a typical simulation. Cell radii vary from 2μm to 15μm.

To account for the results in Fig. 1 (C), we first simulated a mono-disperse system in which all the cells have identical radius (R = 8.5 μm). Because the system crystallizes (Fig. 6 (A) & (B)), we concluded that the dynamics observed in experiments cannot be explained using this model. A 1:1 binary mixture of cells, with different radii gives glass-like behavior for all ϕ, with the relaxation time τα as well as the effective viscosity (defined in Eq. 1) following the VFT behavior (see Appendix B).

Polydispersity and cell-cell interactions

In typical cell tissues, and zebrafish in particular, there is a great dispersion in the cell sizes, which vary in a single tissue by a factor of ~ 5 − 6 [14] (Fig. 1 (B) and Fig. 6 (D)). In addition, the elastic forces characterizing cell-cell interactions are soft, which implies that the cells can penetrate with ηij −(Ri+Rj) < 0 when they are jammed. Thus, both polydispersity as well as soft interactions between cells must control the relaxation dynamics. To test this proposition, we simulated a highly polydisperse system (PDs) in which the cell sizes vary by a factor of ~ 8 (Fig. 1 (D) and Fig. 6 (E)).

A simulation snapshot (Fig. 1 (D)) for ϕ = 0.93 shows that different sized cells are well mixed. The structure of the tissue can be described using the pair correlation function,, where is the number density, δ is the Dirac delta function, is the position of the ηth cell, and the angular bracket ⟨⟩ denotes an average over different ensembles. The g(r) function (Fig. 1 (E)) has a peak around r ~ 17μm, which is approximately the average diameter of the cells. The absence of peaks in g(r) beyond the second one, suggests there is no long range order. Thus, the polydisperse cell system exhibits liquid-like structure even at the high ϕ.

Effective shear viscosity as a function of ϕ

A fit of the experimental data for η using the VFT [24, 25] relation in the range ϕ ≤ 0.87 (Fig. 1 (C)) yields ϕ0 ≈ 0.95 and D ≈ 0.51 [19]. The VFT equation for cells, which is related to the Doolittle equation [26] for fluidity that is based on free space available for motion in an amorphous system [27, 28], is η = η0 exp , where D is the apparent activation energy. In order to compare with experiments, we calculated an effective shear viscosity for the polydisperse system using a Green-Kubo-type relation [29],

The stress-stress tensor Pμν(t) in the above equation is, where μ, ν ∈ (x, y) are the Cartesian components of coordinates, is the force between ith and jth cells and A is the area of the simulation box. Note that should be viewed as a proxy for shear viscosity because it does not contain the kinetic term and the factor is not included in Eqn. (1) because temperature is not a relevant variable in the highly over-damped model for cells considered here.

Plot of as a function of ϕ in Fig. 2 (A), shows qualitatively the same behavior as the estimate of viscosity (using dimensional arguments) made in experiments. Two features about Fig. 1 (C) and Fig. 2 (A) are worth noting. (i) Both simulations and experiments show that up to ϕ ≈ 0.90, follows the VFT relation with ϕ0 ~ 0.94 and D ~ 0.5. More importantly is independent of ϕ when ϕ > 0.90. (ii) The values of ϕ0 and D obtained by fitting the experimental estimate of η to the VFT equation and simulation results are almost identical. Moreover, the onset of the plateau density in simulations and experiments occur at the same value (ϕS ~ 0.90). The overall agreement with experiments is remarkable given that the model was not created to mimic the zebrafish tissue.

To provide additional insights into the dynamics, we calculated the isotropic self-intermediate scattering function, Fs(q, t), where is the wave vector, and is the position of a cell at time t. The degree of dynamic correlation between two cells can be inferred from the decay of Fs(q, t). The angle bracket ⟨…⟩ is an average over different time origins and different trajectories. We chose , where rmax is the position of the first peak in g(r) between all cells (see Fig. 1 (E)). The relaxation time τα is calculated using .

From Figs. 2 (B) & (C), which show Fs(q, t) as a function of t for various ϕ, it is clear that the dynamics become sluggish as ϕ increases. The relaxation profiles exhibit a two step decay with a plateau in the intermediate time scales. The dynamics continues to slow down dramatically until ϕ ≤ 0.90. Surprisingly, the increase in the duration of the plateau in Fs(q, t) ceases when ϕ exceeds ≈ 0.90 (Fig. 2 (C)), a puzzling finding that is also reflected in the dependence of τα on ϕ in Fig. 2 (D). The relaxation time increases dramatically, following the VFT relation, till ϕ ≈ 0.90, and subsequently reaches a plateau (see the inset in Fig. 2 (D)). If the VFT relation continues to hold for all ϕ, as in glasses or in binary mixture of 2D cells (see Appendix B), then the fit yields ϕ0 ≈ 0.95 and D ≈ 0.50. However, the simulations show that τα is nearly a constant when ϕ exceeds 0.90. We should note that the behavior in Fig. 2 (D) differs from the dependence of τα on ϕ for 2D monodisperse polymer rings, used as a model for soft colloids. Simulations [30] showed τα increases till a critical ϕS but it decreases substantially beyond ϕS with no saturation.

Relaxation dynamics of individual cells

Plot of τα as a function of the radius of cells Ri (Fig. 3 (A)) shows nearly eight orders of magnitude change. The size dependence of τα on ϕ is striking. That τα should increase for large sized cells (see the data beyond the vertical dashed line in Fig. 3 (A)) is not unexpected. However, even when cell sizes increase beyond Ri = 4.25 μm, the dispersion in τα is substantial, especially when ϕ exceeds ϕS. The relaxation times for cells with Ri < 4.25 μm are relatively short even though the system as a whole is jammed. For ϕ ≥ 0.90, τα for small sized cells have a weak dependence on ϕ. Although τα for cells with radius < 4 μm is short, it is clear that for a given ϕ (for example ϕ = 0.93) the variations in τα is substantial. In contrast, τα’s for larger cells (R ≥ 7μm) are substantially large, possibly exceeding the typical cell division time in experiments. In what follows, we interpret these results in terms of available free area ⟨Afree⟩ for cells. The smaller sized cells have the largest .

The effect of jamming on the dramatic increase in τα occurs near Ri ≈ 4.5 μm, which is comparable to the length scale of short range interactions. For ϕ ≤ 0.90, τα increases as the cell size increases. However, at higher packing fractions, even cells of similar sizes show substantial variations in τα, which change by almost 3 − 4 orders of magnitude (see the data around the vertical dashed line for ϕ ≥ 0.915 in Fig. 3 (A)). This is a consequence of large variations in the local density (Fig. 14). Some of the similar-sized cells are trapped in the jammed environment, whereas others are in less crowded regions (see Fig. 14). The spread in τα increases dramatically for ϕ > ϕS (≈ 0.90) and effectively overlap with each other. This is vividly illustrated in the the histogram, P(log(τα)) shown in Fig. 3 (B). For ϕ < ϕs, the peak in P(log(τα)) monotonically shifts to higher log(τα) values. In contrast, when ϕ exceeds ϕS there is overlap in P(log(τα)), which is reflected in the saturation of and τα. There are cells (typically with small sizes) that move faster even in a highly jammed environment (see Fig. 12 (C) and Fig. 13). The motions of the fast-moving cells change the local environment, which effectively facilitates the bigger cells to move in a crowded environment (see Fig. 12 (D) and Fig. 13 and and Movie 1 (ϕ = 0.92 > ϕS) and Movie 2 (ϕ = 0.90 = ϕS)). In contrast, for ϕ = 0.85 < ϕS small and large sized cells move without hindrance because of adequate availability of free area (Movie 3). The videos vividly illustrate the large scale facilitated rearrangements that enable the large sized cells to move.

Structure and relaxation behavior for a binary mixture of cells:

(A) A typical simulation snapshot for binary mixture of cells at ϕ = 0.93. (B) The corresponding pair correlation function, g(r), between all the cells. The vertical dashed line is at the first peak position (rmax). (C) Fs(q, t), with , where rmax is the location of the first peak in the g(r), as function of time at various ϕ values. (D) The logarithm of the relaxation time, τα, as a function of ϕ. Over the entire range of ϕ, the increase in τα is well fit by the Vogel-Fucher-Tamman (VFT) relation. Most importantly, the relaxation time does not saturate, which means the evolving tissue cannot be modeled using a 50:50 binary mixture. (E) Effective shear viscosity as a function of ϕ reflects the behavior of τ as a function of ϕ in (D).

Free area decreases monotonically for the binary mixture of cells:

(A) Mean voronoi cell size, ⟨A⟩, as a function of ϕ for the 50:50 binary system. (B) The free area fraction, ϕfree, as a function of ϕ shows that ϕfree decreases monotonically as ϕ increases.

Relaxation in the polydisperse cell system is independent of the waiting time:

(A) Fs(q, t) for ϕ = 0.92 at different waiting times (τω = 106(s)). Regardless of the value of τω, all the Fs(q, t) curves collapse onto a master curve. (B) Relaxation time, ln(τα), as a function of τω. Over a 3 orders of magnitude change in tω, the variation in relaxation times is less than the sample to sample fluctuations, as shown by the error bar.

Measure of ergodicity:

(A) Ergodic measure Ω(t) scaled by the value at t = 0 (Ω(0)) as a function of t for ϕ = 0.85. (B) and (C) Similar plots for ϕ = 0.90 and ϕ = 0.92 respectively. At long time Ω(t)/Ω(0) reach 0.01, 0.016 and 0.026 for ϕ = 0.85, ϕ = 0.90 and ϕ = 0.92 respectively. (D) Ω(0)/Ω(t) as a function of t for ϕ = 0.90. The dashed line shows a linear fit.

Fit of the stress-stress correlation functions to stretched exponential functions:

(A) The stress-stress correlation function ⟨Pμν(t)Pμν(0)⟩ divided by the value at t = 0 Pμν(0)2 as a function of t for ϕ ∈ (0.75 − 0.87). (B) Similar plot for ϕ ∈ (0.89 − 0.93). (C) The long time decay of ⟨Pμν(t)Pμν(0)⟩ is fit to Cs exp , as shown by the dashed lines. The inset shows the dependence of β on ϕ. (D) The data that is fit using the stretched exponential function (black dashed line) is combined with the short time data (blue solid line), which is fit using the cubic spline function. The resulting fits produces a smooth curve⟨Pμν(t)Pμν(0)⟩combined, as shown in the inset.

Cell size dependent structures and dynamics:

(A) Radial distribution function gSS(r) between small sized cells (RS ≤ 4.5μm) at ϕ = 0.905 (blue) and 0.92 (red). These values are greater than ϕs ≈ 0.90 (B) Same as (A) except the results are for gBB(r) between large cells (RB ≥ 12.0μm). (C) Fs(q, t) for cells with RS ≤ 4.5μm at ϕ = 0.905 and ϕ = 0.92. Note that even at these dense packings, the mobility of the smaller sized cells is substantial, which is reflected in the time dependence of Fs(q, t). (D) Fs(q, t) for cells with RB ≥ 12.0μm at ϕ = 0.905 and ϕ = 0.92. The black dashed lines are fits to stretched exponential functions, Fs(q, t) ~ exp where τα is the relaxation time and β is the stretching exponent. The dotted lines correspond to the value .

Simulation snapshot and trajectories for a few smaller and bigger sized cells:

(A) Cells (ϕ = 0.91) are coloured according to their sizes (gray colors). A few small sized cells are shown in different colors (pink, blue, orange, purple, cyan, light purple and yellow). (B) The corresponding trajectories are shown over the entire simulation time. (C) Similar plot as (A) but for a few bigger sized cells shown in purple, yellow, light green, red, cyan and green colors. (D) Same as (B) expect the trajectories of the large sized cells are highlighted. Clearly, the large cells are jammed.

Dynamical rearrangement of jammed cells:

The changing local environment of a randomly selected cell(black) over time. (Top Panels: From left to right t = 9.41τα, 10.01τα and 25.39τα). The black coloured cell is completely jammed by other cells. (Bottom Panels: Form left to right t = 10.97τα, 25.44τα and 27.49τα). Dynamical facilitation, resulting in collective rearrangement of the cells surrounding the black cell, enables it to move in the dynamically created free volume.

The dependence of τα on ϕ for ϕ ≤ ϕS (Fig. 2 (D)) implies that the polydisperse cell systems behave as a soft glass in this regime. On theoretical grounds, it was predicted that P(ln(τα)) ~ exp in glass-forming systems [8]. Remarkably, we found that this prediction is valid in the polydisperse cell system (Fig. 3 (C)). However, above ϕS the predicted relation is not satisfied (see Fig. 3 (D)).

Aging does not explain viscosity saturation

Light scattering experiments found saturation in τα as a function of ϕ at high densities in mono-disperse microgels and silica particles in three dimensions [31]. The plateau in microgels was attributed to “aging” of the samples at high ϕ because it was suspected that the synthetic system falls out of equilibrium. If saturation in viscosity and relaxation time in the embryonic tissue at high ϕ is due to aging then τα should increase sharply as the waiting time, τω, is lengthened. Simulations at different τω with ϕ = 0.92(> ϕS) show (Fig. 9 (A)) that Fs(q, t) is independent of tω as is τα(Fig. 9 (B)). Thus, aging cannot explain the observed viscosity saturation.

Available free area explains viscosity saturation at high ϕ

We explain the saturation in the viscosity by calculating the available free area per cell, as ϕ increases. In a hard disk system, one would expect that the free area would decrease monotonically with ϕ until it is fully jammed at the close packing fraction (~ 0.84 [32, 33]). Because the cells are modeled as soft deformable disks, they could overlap with each other even when fully jammed. Therefore, the region where cells overlap create free area in the immediate neighborhood. The extent of overlap (hij) is reflected in distribution P(hij). The width in P(hij) increases with ϕ, and the peak shifts to higher values of hij (Fig. 4 (A)). The mean, ⟨hij⟩ increases with ϕ (Fig. 4 (B)). Thus, even if the cells are highly jammed at ϕ ≈ ϕS, free area is available because of an increase in the overlap between cells.

When ϕ exceeds ϕS, the mobility of small sized cells facilitates the larger cells to move, as is assumed in the free volume theory of polymer glasses [28, 34–36]. As a result of the motion of small cells, a void is temporarily created, which allows other (possibly large) cells to move. In addition to the release of space, the cells can also interpenetrate (Figs. 4 (A) & (B)). If hij increases, as is the case when the extent of compression increases (Figs. 4 (A)), the available space for nearby cells would also increase. This effect is expected to occur with high probability at ϕS and beyond, resulting in high overlap between the cells. These arguments suggest that the combined effect of polydispersity and cell-cell overlap creates, via the self-propulsion of cells, additional free area that drives larger cells to move even under jammed conditions.

In order to quantify the physical picture given above, we calculated an effective area for each cell by first calculating Voronoi cell area A. A plot for Voronoi tessellation is presented in Fig. 5 (A) for ϕ = 0.93, and the histogram of A is shown in Fig. 5 (B). As ϕ increases, the distribution shifts towards lower Voronoi cell size ⟨A⟩. The mean Voronoi cell size ⟨A⟩ as a function of ϕ in Fig. 5 (C) shows ⟨A⟩ decreases as ϕ is increased. As cells interpenetrate, the Voronoi cell size will be smaller than the actual cell size in many instances (Fig. 5 (A)). To demonstrate this quantitatively, we calculated . The value of Afree could be negative if the overlap between neighbouring cells is substantial; Afree is positive only when the Voronoi cell size is greater than the actual cell size. Positive Afree is an estimate of the available free area. The histograms of Afree for all the packing fractions in Fig. 5 (D) show that the distributions saturate beyond ϕ = 0.90. All the distributions have a substantial region in which Afree is negative. The negative value of Afree increases with increasing ϕ, which implies that the amount of interpenetration between cells increases.

Because of the overlap between the cells, the available free area fraction ϕfree, is higher than the expected free area fraction (1.0 − ϕ) for all ϕ. We define an effective free area fraction ϕfree as, where Np is the number of positive free area in jth snapshots, Nt is the total number of snapshots, Abox is the simulation box area and is the positive free area of ith cell in jth snapshot. The calculated ϕfree, plotted as a function of ϕ in Fig. 5 (E), shows that ϕfree decreases with ϕ until ϕ = 0.90, and then it saturates near a value ϕfree ≈ 0.22 (see the right panel in Fig. 5 (E)). Thus, the saturation in as a function of ϕ is explained by the free area picture, which arises due to combined effect of the size variations and the ability of cells to overlap.

Discussion

Extensive computer simulations of a two-dimensional dense tissue using a particle-based model of soft deformable cells with active self-propulsion have successfully reproduced the dynamical behavior observed in the blastoderm tissue of zebrafish. The dependence of viscosity (η) and relaxation time (τα) (before the saturation) is well fit by the VFT equation. The value of ϕ0 obtained from simulations, ϕ0 ~ 0.95, is close to ϕ0 ~ 0.94 extracted by fitting the experimental data to the VFT equation. Thus, the dynamics for ϕ ≤ ϕS resembles the behavior expected for glass forming systems. Remarkably, the dependence of η on ϕ over the entire range (VFT regime followed by a plateau) may be understood using available free area picture, an idea that was proposed nearly 70 years ago. We discovered that polydispersity as well as the ease of deformation of the cells that creates free area under high jamming conditions, is the mechanism that explains viscosity saturation at high cell densities. The proposed mechanism is an important step that links equilibrium phase transition to dynamics during zebrafish development [37].

We find it surprising that the calculation of viscosity using linear response theory (valid for systems close to equilibrium), and the link to free area quantitatively explain the simulation results and by implication the experimental data for a living and growing tissue. The calculation of free area of the cells is based on the geometrical effects of packing, which in turn is determined by cell to cell contact topology. These considerations, that are firmly established here, explain why equilibrium phase transitions are related to steep increase in viscosity [8] as the area fraction changes over a narrow density range. The absence of aging suggests that, although a large number of cell divisions occur, they must be essentially independent, thus allowing the cells to reach local equilibrium.

eLife Assessment

This fundamental study substantially advances our physical understanding of the sharp increase and saturation of the viscosity of non-confluent tissues with increasing cell density. Through the analysis of a simplified model this study provides compelling evidence that polydispersity in cell size and the softness of cells together can lead to this phenomenon. The work will be of general interest to biologists and biophysicists working on development.

Reviewer #1 (Public Review):

The authors consider data by the Heisenberg group on rheological properties of non-confluent tissue in zebrafish embryos. These data had shown a steep increase and subsequent saturation in viscosity with cell density. The authors introduce a physical agent-based model of such tissues that accounts for the dispersion in cell size and the softness of the cells. The model is inspired by previous models to study glassy dynamics and reveals essential physical features that can explain the observed behavior. It goes beyond previous studies that had analysed the observations in terms of a percolation problem. The numerics are thoroughly done and could have a deep impact on how we describe non-confluent tissues.

A major weakness of the manuscript is the way it is written, which gives the impression to have been done rather carelessly. Several quantities are not properly introduced and at places physical jargon is used that makes the work difficult to access for readers without a background in soft matter.

Reviewer #2 (Public Review):

This paper explores how minimal active matter simulations can model tissue rheology, with applications to the in vivo situation of zebrafish morphogenesis. The authors explore the idea of active noise, particle softness and size heterogeneity cooperating to give rise to surprising features of experimental tissue rheologies (in particular an increase and then a plateau in viscosity with fluid fraction). In general, the paper is interesting from a theoretical standpoint, by providing a bridge between concepts from jamming of particulate systems and experiments in developmental biology. The idea of exploring a free space picture in this context is also interesting.

However, I'm still unsure right now though of how much it can be applied to the specific system that the authors refer to - which could be fixed either by considering other experimental systems/models reported in the recent literature or by doing the following theoretical checks:

- Take your current simulations and smoothly change the ratio of polydispersity from 8 to 0 to see exactly how much dispersity is needed to explain viscosity plateauing, and at which point the transition occurs.

- Cellular self-propulsion does not seem to play a role in zebrafish blastoderm, see Ref. [14]. Active noise has been proposed to play key roles in other systems and you could check whether such active noise could replace self-propulsion in your model, see for example Kim & Campas, Nat Phys, 2021.

- Could you simulate realistic rheological deformations to see how much they match both your expectation and the data?

Reviewer #3 (Public Review):

The authors successfully explain the sharp rise and subsequent saturation of the viscosity in dependence of cell packing fraction in zebrafish blastoderm with the help of a 2d model of soft deformable, polydisperse and self-propelled (active) disks. The main experimental observations can be reproduced and the unusual dependence of the viscosity on packing fraction can be explained by the available free area and the emergent motility of small sized cells facilitating multi-cell rearrangement in a highly jammed environment.

The paper is very well written, the results (experimental as well theoretical) are original and scientifically valid. This is an important contribution to understand rheological properties of non-confluent tissues linking equilibrium and transport properties.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1

As written in my public review I consider the science of this work to be high quality. I have some suggestions for the write-up though. As a general comment, I think that too much has been put into the appendices. In particular, the main text could contain more details about the model.

We are pleased that this Reviewer feels that our work to be of “high quality”. We value the reviewer’s insightful suggestions and comments. Following this Reviewer’s suggestion we have moved certain sections to the main text.

In what follows, we provide responses to each of the reviewer’s inquiry, and indicate the appropriate changes in the revised version.

P2 -

ϕ is introduce as packing fraction - on p3 it’s called cell density. Also it is not clear whether it is an area fraction or a cell number density. Please define properly and I would suggest sticking to one notion.

ϕ is the cell packing fraction. In two dimensions (as is the case in our simulations) it is the area fraction. However, in order to stick to one general notation (independent of dimension) we use “packing fraction” to represent how densely the cells are packed. We changed it the revised manuscript to ensure uniformity.

P3 -

“which should and should slow down the overall dynamics” Typo?

Corrected it in the revised manuscript.

“One would intuitively expect that the ϕfree should decrease with increasing cell density” Please, define ϕfree

ϕfree is defined in Eqn. 4. We ought to have defined it in the introduction.

“When ϕ exceeds ϕS, the free area ϕfree saturates because the soft cells interpenetrate each other,” I suggest clearly distinguishing between biological cells and the agents (disks) used in the simulation. Please, also clarify What interpenetration of agents corresponds to in tissues?

We have rewritten the sentence as, ”The simulations show that when..” Soft disks used in the simulations seem to be not an unrealistic model for biological cells. The small deformations noted in our model is not that different from the cells in the tissues. For visual reference, please see Author response image 1. In the left panel of the figure, a 2D snapshot of the experimental zebrafish tissue, displays the deformation of cells labeled as 1 and 2. Likewise, the right panel illustrates the extent to which such deformations are replicated in the simulation by allowing two cells to overlap (the white area in the right panel of Author response image 1 represents the interpenetration). In the revised manuscript, we have made the necessary change from “soft cells” to “soft disks.”

Snapshots of zebrafish tissue (left panel) (Ref.

[14] main text) and model two dimensional tissue (right). In the right panel the white area represents the overlap and the black vertical line represents the intersection.

“The facilitation mechanism, invoked in glassy systems [22] allows large cells to move with low mobility.” What is the facilitation mechanism?

Facilitation, which is an intuitive idea, that refers to a mechanism by which cells in a in highly jammed environment can only move if the neighboring cells get out of the way. In our case (as shown in the text (Fig.3 (A) and Fig. 13 (A) & (B)) the smaller cells move faster almost independent of ϕ. When a small cell moves, it creates a void which could facilitate neighboring cells (including big ones) to move.

“η (or relaxation time)” I suggest explaining the link between η and the relaxation time.

First, in making this point on aging we only showed that the relaxation time is independent of the waiting time. In the revised manuscript we deleted η.

Although not germane to this study, in the literature on glass transition, it is not uncommon to use relaxation time τα (as a proxy of viscosity η) to describe the dynamics. The relation between τα and η is given by

where G∞ is the “infinite frequency” shear modulus, which holds in unjammed or in liquids. This relation suggests that τα is proportional to η, which is almost never satisfied in glass forming systems.

P5 - “In addition, the elastic forces characterizing cell-cell interactions are soft, which implies that the cells can penetrate with rij − (Ri + Rj) < 0 when they are jammed.” Is this about the model or the biological tissue? Presumably the former, because real cells do not penetrate each other, right? What are rij, Ri and Rj?

This is about the model. The cells are sufficiently soft that they can be deformed, which allows for modest interpenetration. Real cells exhibit similar behavior (see Fig. 1). In inset of Fig. 4 (b) rij is the center to center distance between cells with radii Ri and Rj. It is better to use the word overlap instead of penetrate, which is what we have done in the revised version.

“we simulated a highly polydisperse system (PDs) in which the cell sizes vary by a factor of ∼ 8” Is it important to have a factor 8 - the zebra fish tissue presents a factor 5 − 6?

This is an important question, which is difficult to answer using analytic theory. It does require simulations unfortunately. We do not know a priori the polysipersity value needed to observe saturation in η at high value of ϕ. However, we have shown that the a system with one type of cell (monodisperse) crystallizes. Furthermore, mixtures of two cell types do not show any saturation in η over the parameter range that we explored. A systematic simulation study is needed to explore a range of parameter values to determine the minimum PD, which would match the experimental findings.

We performed 3D simulations to figure out if much less PD would yield saturation in η. Preliminary simulations in three dimensions with a lower value of PD (11.5% with a size variations by a factor of ≈ 2 ) exhibits saturation in the relaxation time. For comparison, the value of PD in the current work is ≈ 24% with a size variation by a factor of 8.

P6 -

“which is related to the Doolittle equation [26] for fluidity ( )” what is the Doolittle equation? Is it important here? Also: “VFT equation for cells”? Is it the same as given on p.2 - so nothing special for cells - or a different one?

Historically, the Doolittle equation was proposed to describe the change in η in terms of free volume in the context polymer systems over 60 years ago. The physics in the polymers is very different from the soft models for cells considered here. Nevertheless, the equations has meaning in the context as well. The Doolittle (other names associated with similar equations are Ferry, Flory... ) equation is given by

, where A and B are constants, V is the total volume and Vhc is the hardcore volume. Essentially, is the relative free volume. It can be shown that one can arrive at the VFT equation starting from the Doolittle equation.

The VFT equation for cells is same as given in page 2, which we restate for completeness. Here, we introduce the apparent activation energy.

“The stress-stress tensor” Why not simply stress tensor?

We have corrected it.

“shows qualitatively the same behavior as the estimate of viscosity (using dimensional arguments) made in experiments.” Where is this shown?

The dependence of viscosity as a function ϕ is shown in Figure 1 (c).

P7 -

Fig 2A caption “dashed line” Maybe full line?

This should be full line. It is fixed in in the revised manuscript.

P8 -

“a puzzling finding that is also reflected” Why is it puzzling?

In figure 2 (C), it shows that the increase in the duration in the plateau of Fs(q,t) ceases when ϕ exceeds ≈ 0.90. This to us is puzzling (always a matter of perspective) because we expected that the duration of Fs(q,t) plateau to increase as a function of ϕ based on the VFT behavior for ϕ ≤ ϕS. As a result, we imagined that the relaxation time τα would continue to increase beyond ϕS. However, the simulations show that the relaxation time is essentially a constant for ϕ > 0.90, which implies that the soft disk system (our model for the tissue) is an unusual with behavior that has no counter part in the material world.

“If the VFT relation continues” –“If the VFT relation continued”

We have fixed it.

First paragraph does not seem to be coherent

What is RS (or Rs)?

RS is the radius of the small cell. In the revised manuscript we have made this clear.

P10 -

Please, define the waiting time.

The waiting time refers to the period between sample preparation and data collection either in experiments or in simulations. In an ergodic system, the properties should not depend on the waiting time provided provided it is large. In other words, after the system reaches thermal equilibrium, the waiting time tω should not have an impact on the properties of the system.

“fully jammed” Please, define.

The term “fully jammed” refers to a state in which the constituent particles in a system do not move. For example, it a hard sphere system at a packing fraction of approximately 0.84 is fully jammed, which implies there is wiggle room for a particle move without violating the excluded volume restriction. At this specific packing fraction, the hard sphere system undergoes a jamming transition, resulting in the particles becoming completely immobile. The nonconfluent tissue modeled here is not fully jammed.

P11 -

Fig.4 it is hard to see that the width of P(hij) increases with ϕ.

Please see Author response image 2 with a less number of curves for a better visualization. We have replaced this figure in the revised version.

Probability of overlap (hij) between two cells, P(hij), for various ϕ values.

“Thus, even if the cells are highly jammed at ϕ ≈ ϕS, free area is available because of an increase in the overlap between cells.” This conclusion seems premature at this point.

The Referee is correct. This is shown in Fig. 5. We amended the ends of the sentence to reflect this observation.

P12 -

“as is the case when the extent of compression increases” extent of compression = density?

This is correct. Extent of compression corresponds to the packing fraction or the density.

“This effect is expected to occur with high probability at ϕS and beyond,” Why? What is special about ϕS.

To achieve high packing fractions beyond a certain value of ϕ soft cells have, which would occur at a certain value ϕS. In the system studied here, ϕ ≈ 0.90 = ϕS. Note that ϕS could be altered by changing the system parameters.

P15 -

“local equilibrium” In a thermodynamic sense? There is also cell migration, so thermodynamic equilibrium does not seem to be appropriate.

This is an important point. The observation that equilibrium concepts hold in what is manifestly a non-equilibrium system is a surprise. It is referred in a thermodynamic sense. We agree with the reviewer because of cell division (in Ref. [14] main text), cell death, thermodynamic equilibrium does not seems to be appropriate. This is exactly the point we raise in the introduction. However, considering the timescale of cell division and death it appears that there may be a local steady state, which we we call a “local equilibrium”. As a consequence phase transition ideas and Green-Kubo relations are applicable. Indeed, a surprise in the conclusion in Ref. [14] is that in the zebrafish morphogenesis equilibrium description seems adequate.

“number of near neighbor cells that is in contact with the ith cell. The jth cell is the nearest neighbor of the ith cell, if hij > 0” A neighbour cell or the nearest neihbor?

A neighbour cell is accurate.

P16 -

“In our model there is no dynamics with only systematic forces because the temperature is zero.” What is a systematic force? I do not understand the sentence.

Systematic force between two cells is defined in Eqn. 5 in the main text. Because temperature is not a relevant variable in our model, we want to emphasize that in the absence of self propulsion, the cells would not move at all.

Reviewer #2

Major comments:

A/ Role of size polydispersity

In the text, and also in the methods (Appendix A), the authors mention that they need large polydispersity of particle sizes to explain the viscous plateau, as the dynamics of small vs large cells are ”dramatically different” (Appendix G). They simulate a system where cell sizes vary by a factor 8, mentioning this is typical in tissues, but I found this quite surprising - this would be heterogeneities in cell volume of 500, many orders of magnitude above what has been measured in tissues. As far as I’m aware, divisions are quite symmetric and synchronous in early vertebrate embryogenesis, so volume variations are expected to be very small (similarly in epithelial tissues, where jamming has been looked at extensively, I’m not aware of examples with ratio of 8 between cell diameters). One question I had is that when the authors look at ”small polydispersity”, there are 50 − 50 mixtures. Would small polydispersity with continuous distributions change this picture? Could they take their current simulations but smoothly change the ratio of polydispersity from 8 to 0 to see exactly how much they need to explain viscosity plateauing, and at which point is the transition?

We thank the reviewer for raising this important question, which was also a concern for Reviewer #1. The value of polydispersity (PD) required to observe such behavior is not known a priori even within the simple model used. We selected a PD value, with a size variation of a factor of 8, guided in part by the experiment (projection onto 2D) shown in Figure 1(B) and Figure 6(D). We also showed that the monodisperse system crystallizes, and the binary system do not show signs of saturation within the explored range of parameter space and ϕ. This suggests that a certain degree of size dispersity is necessary to obtain saturation in η.

As discussed in Appendix B, the binary system is characterized by the variables , where RB and RS represent the radii of the big and small cells, respectively, and the packing fraction ϕ. By more fully exploring the parameter space encompassing λ and ϕ than we did, it maybe possible, as the Referee suggests, that a system with two different cell sizes would yield the experimentally observed dependence of η on ϕ.

As part of an answer to the Reviewer #1 on a the same issue, we mentioned results of preliminary simulations in three dimensions with reduced levels of polydispersity, and discovered that at lower levels of polydispersity (variation in size by a factor of ≈ 2 and polydispersity value 11.50%), the relaxation time does saturate beyond a certain packing fraction (see Fig. 3). We have not established if η, the key quantity of interest, would exhibit a similar behavior in 3D.

(A) τα as a function of ϕ for 11% polydispersity with size variation by a factor of ∼ 2 in the three dimensional system.

(B) Same as (A) except polydispersity value is 24% and a size variation by a factor of ∼ 8.

B/ Role of fluctuations/self-propulsion in this system, and relationship to recent findings

“A priori it is unclear why equilibrium concepts should hold in zebrafish morphogenesis, which one would expect is controlled by non-equilibrium processes such as self-propulsion, growth and cell division. ”

This is raised as a key paradox, but is not very clear to me in the context raised by the authors. In particular, they use self-propulsion as a source of activity and explain the evolution of viscosity but a facilitation process involving re-arrangements/motility. But I don’t think self-propulsion has been argued to play a role in zebrafish blastoderm - Ref 14 argues that this is effectively a zerotemperature phenomenon and that cell motility/rearrangements do not show any correlation with viscosity. So this part of the model assumption was not clear to me in relationship with the proposed experimental system. Active noise has been proposed to play key roles in other systems, including motility-driven and tension fluctuation-driven unjamming (among many others Bi et al, PRX, 2016, Mitchel et al, Nat Comm, 2020, Pinheiro et al, Nat Phys, 2022 as well as Kim & Campas, Nat Physics, 2021) - maybe this is somewhere where the author model could fit? In Kim & Campas, Nat Phys, 2021 in particular, the authors develop simulations of non-confluent tissues with noise, that seems to bear some resemblance to the model developed here, so it would be important to discuss the similarities and distinctions (usually I think polydispersity is not considered indeed). In general, the authors look here at a particle based model, but cells have adhesions with well-defined contact angles, so there is a question of the cross-over between their findings and the large body of recent literature on active foams/vertex models (which are not really discussed there).

We appreciate the lengthy comment here, and there is a lot to unpack. We also thank the referee for the references, some of which we did not know about earlier.

The primary objective of our study is to determine the simplest minimal model that would explain the experimentally observed dependence of viscosity in zebrafish blastoderm tissue as ϕ is increased beyond a certain packing fraction during morphogenesis. In Reference 14, the authors analyzed the data using the framework of rigidity percolation theory and presented evidence of a genuine equilibrium phase transition. Consequently, one would that expect zebrafish blastoderm tissue to be in equilibrium, which is surprising from many perspectives. However, since the tissue is a growing system involving numerous cell divisions and cell death, it is not immediately evident whether the assumption of equilibrium is valid. Indeed, the same problem arises when considering the glass transition where rapid cooling drives the system out of equilibrium. Nevertheless, heat capacity and η are often analyzed using the notion of equilibrium. Hence, considering this issue within the context of our research appears to be reasonable.

To the best of our knowledge, the authors in Ref. 14 did not provide an explanation for the η behavior. The focus was, which was excellent and is the basis on which we initiated this study, was on the use of rigidity percolation theory to explain the results. Indeed, they performed an experiment by mildly reducing myosin II activity, which apparently affects cell motility. The quantitative effect was not reported.

We did not impose any requirement of cell rearrangements etc in the model. There is essentially one variable, free area available, that explains the η dependence on ϕ. It is possible that one can come up with other zero temperature models that could also explain the data. To the best of our knowledge, it has not been proposed.

It would be interesting to set our model in the context of other models that the referee points out. This would be an interesting research topic to explore. The only comment we would like to make is that it is unclear how vertex model for confluent tissues could explain the viscosity data.

C/ Calculation of the effective shear viscosity

The authors calculate viscosity from a Green-Kubo relation, although it would be good to clarify at which time scale (and maybe even shear amplitude) they expect this to be valid. These kinds of model would be expected to show plastic rearrangements for large deformations for instance, could the authors simulate realistic rheological deformations (e.g. Kim & Campas, 2021 applying external shear on the simulations) to see how much this matches both their expectation and the data?

Once it is established that there is local equilibrium (as implied by the use of phase transition ideas to analyse the experimental data in Ref. 14), it is natural to use the Green-Kubo relation to calculate transport properties. Hence, for our purposes, it is valid for all time scales and amplitude. The Reviewer also wonders if the model could be used to simulate response to shear in order to probe rheological properties. There is no conceptual issue here and indeed this is an excellent suggestion that we intend to pursue in the future.

D/ Role of cell adhesion

The authors consider soft elastic disks of different sizes but unless I missed it, there is no adhesion being considered. This is expected to play a key role in jamming and multicellular mechanics, so I think the authors should either look at what this changes in their simulations, or at least discuss why they are neglecting it. One reason I’m asking is that it’s not totally clear to me that the ”free space” picture, coming from the fact that cells can interpenetrate in their model would hold in a model of deformable cells adhering to each other with constant volume (leading to more equilibration of deformations it would seem?).

The referee raises another question regarding the lack of adhesion in the simulations. As pointed out before, we were trying to create a minimal model to account for the experimental observations for η upon changing the packing fraction. Thus, we a coarse-grained model where we considered poly-disperse cells with elastic interactions which recapitulates the experimental observations. The referee is correct that adhesion plays a role in jammed systems, and examination of how it would affect is an aspect that would be interesting to consider in the future. We hasten to add that even systems without attractive adhesion-type interaction become jammed. In principle, in many-body systems, the parameter space is large and one needs to carefully determine which parameter is important for the problem at hand. Therefore, in the first pass we did not find the need to consider the role of adhesion.

Minor comments:

The writing could be condensed in some places, with some details being moved to SI (for instance, section E on ageing is very short and seem more suited for supplements, or at least not as an independent section, note that the figure numbering also jumps to Fig. 9 there, although it’s Fig. 3 just before and Fig. 9 just after - re-ordering into main and supporting figures would be clearer.

We thank the Reviewer for this recommendation. The ageing section, although is short, it does provide a line of evidence that equilibrium approaches could be valid. We have modestly expanded the section by moving Appendix D to the main text, a general suggestion made by Referee 1. We have tried to be consistent in the numbering of figures in the revision.

Reviewer #3

I am very much in favor of the manuscript in its present form - I only suggest commenting (in the manuscript) on the issue described below.

Motivated by the fact that the experimental system consists of living, motile cells the authors use an active particle model (eq. 6) with stochastic selfpropulsion as the only source for noise (zero-temperature). It would be useful to elaborate briefly how important this stochastic self-propulsion is for the emergent rheological properties of the system (as summarized above): would these properties also be present in the “passive” version of the same model at “non-vanishing” temperature, and if not, why? Or analogously in a “passive” version which is “shaken”, reminiscent of shaken granular matter? To clarify these issues would relate this study to (or discriminate it from) passive, but complex, liquids or granular matter.

We appreciate the reviewer’s positive feedback on our work. The reviewer has raised an important question concerning our model in which self-propulsion serves as the source of noise. Without self-propulsion, the system would come to a stationary state after reaching mechanical equilibrium. As mentioned in Eqn. (6) (in the main text), we can define a characteristic time . It is possible that scaling the time t by τ would not alter the results.

The second question raised by the reviewer is also important. A passive version of the model would be to consider Eq. 6 in our article, and instead of using activity use the standard stochastic force. The resulting force would be at a finite temperature,. The coefficient of noise (a diffusion term) would be related to γi through the Fluctuation dissipation theorem(FDT)). Such a system of equations cannot ne mapped to Eq. 6 in which µ and γi are independently varied. It is unlikely that such a model, incorporating a “non-vanishing” temperature, would not result in the observed dependence of η on ϕ for the following reason. The passive model represents a polydisperse system, which would form a glass with η increasing with volume fraction, following the VFT law, as has been demonstrated in the glass transition literature for harmonic glasses. The other proposal whether the shaken version version would explain the experiments is also interesting. These are worth pursuing in future studies.

Version 2

Abstract

A recent experiment on zebrafish blastoderm morphogenesis showed that the viscosity (η) of a non-confluent embryonic tissue grows sharply until a critical cell packing fraction (ϕS). The increase in η up to ϕS is similar to the behavior observed in several glass forming materials, which suggests that the cell dynamics is sluggish or glass-like. Surprisingly, η is a constant above ϕS. To determine the mechanism of this unusual dependence of η on ϕ, we performed extensive simulations using an agent-based model of a dense non-confluent two-dimensional tissue. We show that polydispersity in the cell size, and the propensity of the cells to deform, results in the saturation of the available free area per cell beyond a critical packing fraction. Saturation in the free space not only explains the viscosity plateau above ϕS but also provides a relationship between equilibrium geometrical packing to the dramatic increase in the relaxation dynamics.

Introduction

There is great interest in characterizing the mechanical and dynamical properties of embryonic tissues because they regulate embryo development [1–5]. Measurements of bulk properties, such as viscosity and elastic modulus, and the dynamics of individual cells through imaging techniques, have been interpreted by adapting concepts developed to describe phase transitions, glass transition, and active matter [6–9]. Several experiments have shown that during embryo morphogenesis, material properties of the tissues change dramatically [10– 14]. Of relevance to our study is a remarkable finding that provided evidence that a phase transition (PT) occurs during zebrafish blastoderm morphogenesis, which was analyzed using rigidity percolation theory [14–17]. The authors also estimated the viscosity (η) of the blastoderm tissue using the micropipette aspiration technique [13, 18]. It was found that change in η is correlated with cell connectivity (⟨C⟩), rising sharply over a narrow range of ⟨C⟩. Surprisingly, a single geometrical quantity, the cell-cell contact topology controls both the rigidity PT as well as changes in η in this non-confluent tissue, thus linking equilibrium and transport properties.

Here, we focus on two pertinent questions that arise from the experiments on zebrafish blastoderm. First, experiments [14] showed that η increases as a function of the cell packing fraction (ϕ) till ϕ ≤ 0.87. The dependence of η on ϕ follows the well-known Vogel-FulcherTamman (VFT) law [19], which predicts that η grows monotonically with ϕ. The VFT law, which is commonly used to analyze the viscosity of a class of glass forming materials [20], is given by where ϕ is a constant. Surprisingly, for packing fractions, ϕ ≥ ϕS ≈ 0.90, η deviates from the VFT law and is independent of ϕ, which cannot be explained using conventional theories for glasses [8, 21]. Second, the experimental data [14] was interpreted using equilibrium rigidity percolation theory [15–17] for an embryonic tissue in which cells undergo random cell divisions. A priori it is unclear why equilibrium concepts should hold in zebrafish morphogenesis, which one would expect is controlled by non-equilibrium processes such as self-propulsion, growth and cell division.

We show that the two conundrums (saturation of η at high packing fractions and the use of equilibrium statistical mechanics in a growing system to explain phase transition) may be rationalized by (i) assuming that the interactions between the cells are soft, (ii) the cell sizes are highly heterogeneous (polydisperse), which is the case in zebrafish blastoderm. Using an agent-based (particle) simulation model of a two-dimensional non-confluent tissue, we explore the consequences of varying ϕ (see section Materials and Methods for the definition) of interacting self-propelled polydisperse soft cells, on η. The central results of our study are: (i) The calculated effective viscosity (for details see Appendix F), for the polydisperse cell system, shows that for ϕ ≤ ϕS ≈ 0.90 the increase in viscosity follows the VFT law. Just as in experiments, η is essentially independent of ϕ at high (≥ ϕS) packing fractions. The unusual dependence of η at ϕ ≥ ϕS is quantitatively explained using the notion of available free area fraction (ϕfree), which is the net void space that can be explored by the cells when they are jammed. At high densities, a given cell requires free space in order to move. The free area is created by movement of the neighboring cells into the available void space. One would intuitively expect that the ϕfree should decrease with increasing packing fractions, due to cell jamming, which should slow down the overall dynamics. Indeed, we find that ϕfree decreases with increasing packing fraction (ϕ) until ϕS. The simulations show that when ϕ exceeds ϕS, the free area ϕfree saturates because the soft cells (characterized by “soft deformable disks”) can overlap with each other, resulting in the collective dynamics of cells becoming independent of ϕ for ϕ ≥ ϕS. As a consequence η saturates at high ϕ. (iii) Cells whose sizes are comparable to the available free area move almost like a particle in a liquid. The motility of small sized cells facilitates adjacent cells to move through multi-cell rearrangement even in a highly jammed environment. The facilitation mechanism, invoked in glassy systems [22] allows large cells to move with low mobility. A cascade of such facilitation processes enable all the cells to remain dynamic even above the onset packing fraction of the phase transition. (iv) We find that the relaxation time does not depend on the waiting time for measurements even in the regime where viscosity saturates. In other words, there is no evidence of aging even in the regime where viscosity saturates. Strikingly, the tissue exhibits ergodic [23] behavior at all densities. The cell-based simulations, which reproduce the salient experimental features, may be described using equilibrium statistical mechanics, thus providing credence to the use of cell contact mechanics to describe both rigidity PT and dynamics in an evolving non-confluent tissue [14].

Results

Experimental results

We first describe the experimental observations, which serve as the basis for carrying out the agent-based simulations. Fig. 1(A) shows the bright-field images of distinct stages during zebrafish morphogenesis. A 2D section of zebrafish blastoderm (Fig. 1 (B)) shows that there is considerable dispersion in cell sizes. The statistical properties of the cell sizes are shown in Fig. 8 (D). Fig. 1 (C) shows that η increases sharply over a narrow ϕ range, and saturates when ϕ exceeds ϕS ≈ 0.90.

Structure and viscosity of non-confluent tissues:

(A) Bright-field single-plane images of an exemplary embryo of zebrafish before (t = −60 min), at the onset (t = 0 min), and after blastoderm spreading (t = 60 min). (B) Snapshot of 2D confocal sections at the 1st–2nd deep-cell layer of the blastoderm at t = 60 min. (A) and (B) are taken from [14]. (C) Viscosity η of zebrafish blastoderm as a function of ϕ in a log-linear scale using the data from [14]. The dashed line is the fit to VFT equation. Note that η does not change significantly beyond ϕ ≥ 0.87. (D) A typical snapshot taken from cell-based simulations for ϕ = 0.93. Cells are colored according to their radii (in μm) (color bar shown on the right). (E) The pair correlation function, g(r), as a function of r for ϕ = 0.93. The vertical dashed line is the position of the first peak (rmax = 17.0 μm). The pair correlation function does not exhibit signs of long-range order. Scale bars in (A) is 100 μm and in (B) is 50 μm.

To account for the results in Fig. 1 (C), we first simulated a mono-disperse system in which all the cells have identical radius (R = 8.5 µm). Because the system crystallizes (Fig. 8 (A) & (B)), we concluded that the dynamics observed in experiments cannot be explained using this model. A 1:1 binary mixture of cells, with different radii gives glass-like behavior for all ϕ, with the relaxation time τα as well as the effective viscosity (defined in Eq. 1) following the VFT behavior (see Appendix B).

Polydispersity and cell-cell interactions

In typical cell tissues, and zebrafish in particular, there is a great dispersion in the cell sizes, which vary in a single tissue by a factor of ∼ 5 − 6 [14] (Fig. 1 (B) and Fig. 8 (D)). In addition, the elastic forces characterizing cell-cell interactions are soft, which implies that the cells can overlap, with rij −(Ri +Rj) < 0 when they are jammed ((Fig. 1 (B) and (Fig. 1 (D)). Thus, both polydispersity as well as soft interactions between the cells must control the relaxation dynamics. To test this proposition, we simulated a highly polydisperse system (PDs) in which the cell sizes vary by a factor of ∼ 8 (Fig. 1 (D) and Fig. 8 (E)).

A simulation snapshot (Fig. 1 (D)) for ϕ = 0.93 shows that different sized cells are well mixed. In other words, the cells do not phase separate. The structure of the tissue can be described using the pair correlation function, , where is the number density, δ is the Dirac delta function, is the position of the ith cell, and the angular bracket ⟨⟩ denotes an average over different ensembles. The g(r) function (Fig. 1 (E)) has a peak around r ∼ 17µm, which is approximately the average diameter of the cells. The absence of peaks in g(r) beyond the second one, suggests there is no long range order. Thus, the polydisperse cell system exhibits liquid-like structure even at the high ϕ.

Effective shear viscosity as a function of ϕ

A fit of the experimental data for η using the VFT [24, 25] relation in the range ϕ ≤ 0.87 (Fig. 1 (C)) yields ϕ0 ≈ 0.95 and D ≈ 0.51 [19]. The VFT equation for cells, which is related to the Doolittle equation [26] for fluidity that is based on free space available for motion in an amorphous system [27, 28], is , where D is the apparent activation energy. In order to compare with experiments, we calculated an effective shear viscosity for the polydisperse system using a Green-Kubo-type relation [29], The stress tensor Pµν(t) in the above equation is, where µ, ν ∈ (x, y) are the Cartesian components of coordinates, is the force between ith and jth cells and A is the area of the simulation box. Note that should be viewed as a proxy for shear viscosity because it does not contain the kinetic term and the factor is not included in Eqn. (1) because temperature is not a relevant variable in the highly over-damped model for cells considered here.

Plot of as a function of ϕ in Fig. 2 (A), shows qualitatively the same behavior as the estimate of viscosity (using dimensional arguments) made in experiments. Two features about Fig. 1 (C) and Fig. 2 (A) are worth noting. (i) Both simulations and experiments show that up to follows the VFT relation with ϕ0 ∼ 0.94 and D ∼ 0.5. More importantly is independent of ϕ when ϕ > 0.90. (ii) The values of ϕ0 and D obtained by fitting the experimental estimate of η to the VFT equation and simulation results are almost identical. Moreover, the onset of the plateau packing fraction in simulations and experiments occur at the same value (ϕS ∼ 0.90). The overall agreement with experiments is remarkable given that the model was not created to mimic the zebrafish tissue.

Saturation in viscosity and relaxation time:

(A) Effective viscosity as a function of ϕ, with the solid line being the fit to VFT equation. The inset shows at high ϕ. The dashed line in the inset is the expected behavior assuming that the VFT relation holds at all ϕ. (B) The self-intermediate scattering function Fs(q, t) as a function of t for 0.70 ≤ ϕ ≤ 0.905. The dashed line corresponds to . (C) A similar plot for ϕ > 0.905. (D) The logarithm of the relaxation time τα(s) as a function of ϕ. The VFT fit is given by the dashed line. The inset shows a zoomed-in view for ϕ ≥ ϕS.

To provide additional insights into the dynamics, we calculated the isotropic selfintermediate scattering function, Fs(q, t), where is the wave vector, and is the position of a cell at time t. The degree of dynamic correlation between two cells can be inferred from the decay of Fs(q, t). The angle bracket ⟨…⟩ is an average over different time origins and different trajectories. We chose , where rmax is the position of the first peak in g(r) between all cells (see Fig. 1 (E)). The relaxation time τα is calculated using .

From Figs. 2 (B) & (C), which show Fs(q, t) as a function of t for various ϕ, it is clear that the dynamics become sluggish as ϕ increases. The relaxation profiles exhibit a two step decay with a plateau in the intermediate time scales. The dynamics continues to slow down dramatically until ϕ ≤ 0.90. Surprisingly, the increase in the duration of the plateau in Fs(q, t) ceases when ϕ exceeds ≈ 0.90 (Fig. 2 (C)), a puzzling finding that is also reflected in the dependence of τα on ϕ in Fig. 2 (D). The relaxation time increases dramatically, following the VFT relation, till ϕ ≈ 0.90, and subsequently reaches a plateau (see the inset in Fig. 2 (D)). If the VFT relation continued to hold for all ϕ, as in glasses or in binary mixture of 2D cells (see Appendix B), then the fit yields ϕ0 ≈ 0.95 and D ≈ 0.50. However, the simulations show that τα is nearly a constant when ϕ exceeds 0.90. We should note that the behavior in Fig. 2 (D) differs from the dependence of τα on ϕ for 2D monodisperse polymer rings, used as a model for soft colloids. Simulations [30] showed τα increases till a critical ϕS but it decreases substantially beyond ϕS with no saturation.

Relaxation dynamics of individual cells

Plot of τα as a function of the radius of cells Ri (Fig. 3 (A)) shows nearly eight orders of magnitude change. The size dependence of τα on ϕ is striking. That τα should increase for large sized cells (see the data beyond the vertical dashed line in Fig. 3 (A)) is not unexpected. However, even when cell sizes increase beyond Ri = 4.25 µm, the dispersion in τα is substantial, especially when ϕ exceeds ϕS. The relaxation times for cells with Ri < 4.25 µm are relatively short even though the system as a whole is jammed. For ϕ ≥ 0.90, τα for small sized cells have a weak dependence on ϕ. Although τα for cells with radius < 4 µm is short, it is clear that for a given ϕ (for example ϕ = 0.93) the variations in τα is substantial. In contrast, τα’s for larger cells (R ≥ 7µm) are substantially large, possibly exceeding the typical cell division time in experiments. In what follows, we interpret these results in terms of available free area ⟨Afree⟩ for cells. The smaller sized cells have the largest (RS is the radius of the small cell).

Spectrum of relaxation times:

(A) Scatter plot of relaxation times τα(s) as a function of cell radius. From top to bottom, the plot corresponds to decreasing ϕ. The vertical dashed line is for Ri = 4.25 μm, beyond which the τα changes sharply at high packing fractions. (B) Histogram P (ln(τα)) as a function of ln(τα). Beyond ϕ = 0.90 (ϕS), the histogram peaks do not shift substantially towards a high τα values. (C) For ϕ ≤ ϕS P (ln(τα)) (scaled by P max(ln(τα))) falls on a master curve, as described in the main text. (D) Same as (C) except the results are for ϕ > 0.90. The data deviates from the Gaussian fit, shown by the dashed line.

The effect of jamming on the dramatic increase in τα occurs near Ri ≈ 4.5 µm, which is comparable to the length scale of short range interactions. For ϕ ≤ 0.90, τα increases as the cell size increases. However, at higher packing fractions, even cells of similar sizes show substantial variations in τα, which change by almost 3 − 4 orders of magnitude (see the data around the vertical dashed line for ϕ ≥ 0.915 in Fig. 3 (A)). This is a consequence of large variations in the local density (Fig. 14). Some of the similar-sized cells are trapped in the jammed environment, whereas others are in less crowded regions (see Fig. 14). The spread in τα increases dramatically for ϕ > ϕS (≈ 0.90) and effectively overlap with each other. This is vividly illustrated in the the histogram, P(log(τα)) shown in Fig. 3 (B). For ϕ < ϕs, the peak in P(log(τα)) monotonically shifts to higher log(τα) values. In contrast, when ϕ exceeds ϕS there is overlap in P(log(τα)), which is reflected in the saturation of and τα. There are cells (typically with small sizes) that move faster even in a highly jammed environment (see Fig. 12 (C) and Fig. 13). The motions of the fast-moving cells change the local environment, which effectively facilitates the bigger cells to move in a crowded environment (see Fig. 12 (D) and Fig. 13 and and Movie 1 (ϕ = 0.92 > ϕS) and Movie 2 (ϕ = 0.90 = ϕS)). In contrast, for ϕ = 0.85 < ϕS small and large sized cells move without hindrance because of adequate availability of free area (Movie 3). The videos vividly illustrate the large scale facilitated rearrangements that enable the large sized cells to move.

The dependence of τα on ϕ for ϕ ≤ ϕS (Fig. 2 (D)) implies that the polydisperse cell systems behave as a soft glass in this regime. On theoretical grounds, it was predicted that in glass-forming systems [8]. Remarkably, we found that this prediction is valid in the polydisperse cell system (Fig. 3 (C)). However, above ϕS the predicted relation is not satisfied (see Fig. 3 (D)).

Available free area explains viscosity saturation at high ϕ

We explain the saturation in the viscosity by calculating the available free area per cell, as ϕ increases. In a hard disk system, one would expect that the free area would decrease monotonically with ϕ until it is fully jammed at the close packing fraction (∼ 0.84 [31, 32]). Because the cells are modeled as soft deformable disks, they could overlap with each other even when fully jammed. Therefore, the region where cells overlap create free area in the immediate neighborhood. The extent of overlap (hij) is reflected in distribution P(hij). The width in P(hij) increases with ϕ, and the peak shifts to higher values of hij (Fig. 4 (A)). The mean, ⟨hij⟩ increases with ϕ (Fig. 4 (B)). Thus, even if the cells are highly jammed at ϕ ≈ ϕS, free area is available because of an increase in the overlap between cells (see Fig. 5). When ϕ exceeds ϕS, the mobility of small sized cells facilitates the larger cells to move, as is assumed in the free volume theory of polymer glasses [28, 33–35]. As a result of the motion of small cells, a void is temporarily created, which allows other (possibly large) cells to move. In addition to the release of space, the cells can also interpenetrate (Figs. 4 (A) & (B)). If hij increases, as is the case when the extent of compression increases (Figs. 4 (A)), the available space for nearby cells would also increase. This effect is expected to occur with high probability at ϕS and beyond, resulting in high overlap between the cells. These arguments suggest that the combined effect of polydispersity and cell-cell overlap creates, via the self-propulsion of cells, additional free area that drives larger cells to move even under jammed conditions.

Density dependent cell-cell overlap :

(A) Probability of overlap (hij) between two cells, P (hij), for various ϕ values. The peak in the distribution function shifts to higher values as ϕ increases. (B) Mean ⟨hij⟩ = dhijP (hij) as a function of ϕ. Inset shows a pictorial illustration of h12 between two cells with radii R1 and R2 at a distance r12.

Changes in free area fraction with ϕ:

(A) Voronoi tessellation of cells for ϕ = 0.93 for a single realization. The orange circles represent actual cell sizes. The blue polygons show the Voronoi cell size. (B) Distribution of Voronoi cell size A as a function of ϕ. (C) Mean Voronoi cell size ⟨A⟩ as a function of ϕ. A zoomed-in view for ϕ > 0.860 is shown in the inset. (D) Distribution of free area P (Afree) for all ϕ. The vertical blue dashed line shows that the maximum in the distribution is at Afree ∼ 50μm2. (E) Free area fraction ϕfree as a function of ϕ. Note that ϕfree saturates beyond ϕ = 0.90. An expanded view of the saturated region is shown in the right panel of (E).

In order to quantify the physical picture given above, we calculated an effective area for each cell by first calculating Voronoi cell area A. A plot for Voronoi tessellation is presented in Fig. 5 (A) for ϕ = 0.93, and the histogram of A is shown in Fig. 5 (B). As ϕ increases, the distribution shifts towards lower Voronoi cell size ⟨A⟩. The mean Voronoi cell size ⟨A⟩ as a function of ϕ in Fig. 5 (C) shows ⟨A⟩ decreases as ϕ is increased. As cells interpenetrate, the Voronoi cell size will be smaller than the actual cell size in many instances (Fig. 5 (A)). To demonstrate this quantitatively, we calculated . The value of Afree could be negative if the overlap between neighbouring cells is substantial; Afree is positive only when the Voronoi cell size is greater than the actual cell size. Positive Afree is an estimate of the available free area. The histograms of Afree for all the packing fractions in Fig. 5 (D) show that the distributions saturate beyond ϕ = 0.90. All the distributions have a substantial region in which Afree is negative. The negative value of Afree increases with increasing ϕ, which implies that the amount of interpenetration between cells increases.

Because of the overlap between the cells, the available free area fraction ϕfree, is higher than the expected free area fraction (1.0 − ϕ) for all ϕ. We define an effective free area fraction ϕfree as, where Np is the number of positive free area in jth snapshots, Nt is the total number of snapshots, Abox is the simulation box area and is the positive free area of ith cell in jth snapshot. The calculated ϕfree, plotted as a function of ϕ in Fig. 5 (E), shows that ϕfree decreases with ϕ until ϕ = 0.90, and then it saturates near a value ϕfree ≈ 0.22 (see the right panel in Fig. 5 (E)). Thus, the saturation in as a function of ϕ is explained by the free area picture, which arises due to combined effect of the size variations and the ability of cells to overlap.

F. Aging does not explain viscosity saturation

Our main result, which we explain by adopting the free volume theory developed in the context of glasses [28, 33–35], is that above a critical packing fraction ϕS ∼ 0.90 the viscosity saturates. Relaxation time, τα, measured using dynamic light scattering, in nearly monodisperse microgel poly(Nisopropylacrylamide) (PNiPAM) [36] was found to depend only weakly on the volume fraction (3D), if ϕV, exceeds a critical value. It was suggested that the near saturation of τα at high ϕV is due to aging, which is a non-equilibrium effect. If saturation in viscosity and relaxation time in the embryonic tissue at high ϕ is due to aging then τα should increase sharply as the waiting time, τω, is lengthened. We wondered if aging could explain the observed saturation of η in the embryonic tissue above ϕS. If aging causes the plateau in the tissue dynamics, then η or τα should be an increasing function of the waiting time, τω. To test the effect of τω on τα, we calculated the self-intermediate scattering function Fs(q, t+τω) as a function of t by varying τω over three orders of magnitude at ϕ = 0.92 (Fig. 6 (A)). There is literally no change in Fs(q, t + τω) over the entire range of tω. We conclude that, τα, extracted from Fs(q, t+τω) is independent of τω. The variations in τα (Fig. 6 (B)), with respect to τω, is significantly smaller than the errors in the simulation. Thus, the saturation in η or τα when ϕ > ϕS is not a consequence of aging.

Relaxation in the polydisperse cell system is independent of the waiting time:

(A) Fs(q, t) for ϕ = 0.92 at different waiting times (τω = 106(s)). Regardless of the value of τω, all the Fs(q, t) curves collapse onto a master curve. (B) Relaxation time, ln(τα), as a function of τω. Over a 3 orders of magnitude change in tω, the variation in relaxation times is less than the sample to sample fluctuations, as shown by the error bar.

There are two implications related to the absence of aging in the dynamics of the nonconfluent embryonic tissues. (i) Although active forces drive the dynamics of the cells, as they presumably do in reality, the cell collectives can be treated as being near equilibrium, justifying the use of Green-Kubo relation to calculate η. (ii) Parenthetically, we note that the absence of significant non-equilibrium effects, even though Zebrafish is a living system, further justifies the use of equilibrium rigidity percolation theory to analyze the experimental data [14].

Discussion

Extensive computer simulations of a two-dimensional dense tissue using a particle-based model of soft deformable cells with active self-propulsion have successfully reproduced the dynamical behavior observed in the blastoderm tissue of zebrafish. The dependence of viscosity (η) and relaxation time (τα) (before the saturation) is well fit by the VFT equation. The value of ϕ0 obtained from simulations, ϕ0 ∼ 0.95, is close to ϕ0 ∼ 0.94 extracted by fitting the experimental data to the VFT equation. Thus, the dynamics for ϕ ≤ ϕS resembles the behavior expected for glass forming systems. Remarkably, the dependence of η on ϕ over the entire range (VFT regime followed by a plateau) may be understood using available free area picture with essentially a single parameter, an idea that was proposed nearly 70 years ago. We discovered that polydispersity as well as the ease of deformation of the cells that creates free area under high jamming conditions, is the mechanism that explains viscosity saturation at high cell densities. The mechanism suggested here is an important step that links equilibrium phase transition to dynamics during zebrafish development [37].

One could legitimately wonder if the extent of polydispersity (PD) used in the soft discs simulations, which seems substantial, is needed to recapitulate the observed dependence of η on ϕ. Furthermore, such large values of PD may not represent biological tissues. Although the choice of PD was made in part by the two-dimensional projection of area reported in experiments [14], it is expected that PD values have to be less in three dimensions. We performed preliminary simulations in three dimensions with considerably reduced PD, and calculated the dependence of relaxation time (τα) as a function of ϕ. The results show that τα does indeed saturate at high volume fractions.

The proposed model neglects adhesive interactions between cells, which of course is not unimportant. It is crucial to wonder if the proposed mechanism would change if cell-cell adhesion is taken into account. We wanted to create the simplest model to explain the experimental data. We do think that realistic values of adhesion strength does not significantly alter the forces between cells [38]. Thus, we expect a similar mechanism. Furthermore, the physics of the dynamics in glass forming materials does not change in systems with and without attractive forces [8]. For these reasons, in the first pass we neglected adhesion, whose effects have to be investigated in the future.

In the physical considerations leading to Eq. 6, the random activity term (µ) plays an important role. Is it possible to create a passive model by maintaining the system at a finite temperature using stochastic noise with µ = 0, which would show the observed viscosity behavior? First, in such a system of stochastic equations, the coefficient of noise (a diffusion constant) would be related to γi in Eq.6 through fluctuation dissipation theorem (FDT). Thus, only γi can be varied. In contrast, in Eq.6 the two parameters (γi and µ) maybe independently changed, which implies that the two sets of stochastic equations of motion are not equivalent. Second, the passive system describes particles that interact by soft Hertz potential. In analogy with systems in which the particles interact with harmonic potential [39], we expect that the passive model would form a glass in which the viscosity would follow the VFT law.

We find it surprising that the calculation of viscosity using linear response theory (valid for systems close to equilibrium), and the link to free area quantitatively explain the simulation results and by implication the experimental data for a living and growing tissue. The calculation of free area of the cells is based on the geometrical effects of packing, which in turn is determined by cell to cell contact topology. These considerations, that are firmly established here, explain why equilibrium phase transitions are related to steep increase in viscosity [8] as the packing fraction changes over a narrow range. The absence of aging suggests that, although a large number of cell divisions occur, they must be essentially independent, thus allowing the cells to reach local equilibrium.

eLife Assessment

This fundamental study substantially advances our physical understanding of the sharp increase and saturation of the viscosity of non-confluent tissues with increasing cell density. Through the analysis of a simplified model this study provides compelling evidence that polydispersity in cell size and the softness of cells together can lead to this phenomenon. The work will be of general interest to biologists and biophysicists working on development.

Reviewer #1 (Public Review):

The authors consider data by the Heisenberg group on rheological properties of non-confluent tissue in zebrafish embryos. These data had shown a steep increase and subsequent saturation in viscosity with cell density. The authors introduce a physical agent-based model of such tissues that accounts for the dispersion in cell size and the softness of the cells. The model is inspired by previous models to study glassy dynamics and reveals essential physical features that can explain the observed behavior. It goes beyond previous studies that had analysed the observations in terms of a percolation problem. The numerics is thoroughly done and could have a deep impact on how we describe non-confluent tissues.

Reviewer #2 (Public Review):

This paper explores how minimal active matter simulations can model tissue rheology, with applications to the in vivo situation of zebrafish morphogenesis. The authors explore the idea of active noise, particle softness and size heterogeneity cooperating to give rise to surprising features of experimental tissue rheologies (in particular an increase and then a plateau in viscosity with fluid fraction). In general, the paper is interesting from a theoretical standpoint, by providing a bridge between concepts from jamming of particulate systems and experiments in developmental biology. The idea of exploring a free space picture in this context is also interesting. However, I'm still unsure right now though of how much it can be applied to the specific system that the authors refer to - which could be fixed either by doing theoretical checks or considering other experimental systems/models reported in the recent literature.

Reviewer #3 (Public Review):

The authors successfully explain the sharp rise and subsequent saturation of the viscosity in dependence of cell packing fraction in zebrafish blastoderm with the help of a 2D model of soft deformable, polydisperse and self-propelled (active) disks. The main experimental observations can be reproduced and the unusual dependence of the viscosity on packing fraction can be explained by the available free area and the emergent motility of small sized cells facilitating multi-cell rearrangement in a highly jammed environment.

The paper is very well written, the results (experimental as well as theoretical) are original and scientifically valid. This is an important contribution to understanding the rheological properties of non-confluent tissues linking equilibrium and transport properties.

Author Response

The following is the authors’ response to the previous reviews.

We are pleased that Reviewers 1 and 3 have recommended that the revised paper be published.

Reviewer #2

For point A: Their preliminary simulation in 3D looks also nice, although it’s referenced in the discussion but not actually included in manuscript - I would advise adding it even under the mention of preliminary.

We appreciate the reviewer for liking our 3D results and suggesting to include them in the manuscript. However, these are preliminary results of our ongoing work. We are yet to establish the corresponding viscosity results quantitatively in the 3D simulations. Because the relationship between viscosity and relaxation time is not (always) linear in glass forming systems, we hesitate to report our results for publication. We hope to report the new results as part of a separate work.

For point B/C: I see some of the points of the authors - although not all of it made it in the main text. I still have some points that puzzle me. For instance, the authors mention that a single value of viscosity (from Green-Kubo) is ”valid for all time scales and amplitude”. This sounds very surprising to me for a complex fluid even at equilibrium: doesn’t it for instance assume linear response (hence small amplitudes)? Fast vs slow probing of a complex medium should also matter (see refs previously mentioned). Related to this, it’s not clear how can self-propulsion not matter if one would shear the system at a finite time scale, given past work on motility-driven unjamming and the mechanism of the authors from facilitation ( wouldn’t shearing at time scales larger vs smaller than the typical time for given cells to spontaneously rearrange from self-propulsion change drastically the effective complex modulus of the system?)

There might be a slight misunderstanding between the reviewer and us when

we say ‘single value of viscosity is valid for all time-scales and amplitude’. Let us explain this point more carefully. In our problem, we are studying the dynamics of a many body system which is undergoing Brownian dynamics where the fluctuation-dissipation theorem need not be valid (as the friction and the selfpropulsion noise strength are not related via Fluctuation-Dissipation Theorem). Now, for us to use the concepts of linear-response (which in the present study are the Green-Kubo relations for the transport coefficients in terms of timecorrelations functions), we need to show that the within the simulation time, the system has reached state that could be described using an “equilibrium” probability measure. This is the precise reason we calculated the ergodicity measure, which is a way to show that all the phase-space have been sampled uniformly under the given Brownian dynamics. This suggests (does not prove) that the system has attained a stationary probability measure (i.e, near equilibrium) for the value of self-propulsion used. Now for this value of self-propulsion, the Green-Kubo relations hold for ‘any time-scale of the simulations’ so that we can perform a time average over the trajectories of the particles (which is an alias of the stationary probability measure under the values of self-propulsion used). If we change the amplitude of the self-propulsion, we need to again compute the ergodicity measure and show the stationarity of the probability measure. If the system is ergodic with respect to the new self-propulsion, we can again use Green-Kubo for the simulations. Note that we will definitely get a different value of viscosity under the new self-propulsion as the shear-stresses generated will be different but the Green-Kubo holds. If the system is not ergodic, for the self-propulsion with the new amplitude, we cannot use Green-Kubo relations. Also a priori, one cannot say what is a large/small amplitude of self-propulsion because it has to be compared with the intrinsic energy scale, which is encoded in the energy function, which is difficult to say without explicit calculations.

This is what we meant when we said, ‘single value of viscosity is valid for all time-scales and amplitude’. It is valid for time-scales of the simulations for a given amplitude of self-propulsion only if the system is ergodic. Note that if the system is not ergodic, then the results of Ref. [14] (in the main text) could be questioned on theoretical grounds, because they were analyzed using 3 the equilibrium rigidity percolation theory. Nevertheless, the authors of Ref. [14] showed that equilibrium phase transition theory works in tissues. For these reasons, we have been, just like the Reviewer, puzzled that equilibrium ideas appear to be valid in the cell system. Additional theoretical work has to be done to clarify these links in tissues. Although this is not the last word, we hope this clarifies our view point.

For point D: I agree with the simplicity argument, although the added sentence from the discussion “Furthermore, the physics of the dynamics in glass forming materials does not change in systems with and without attractive forces” seems a bit strong given works like Lois et al., PRL, 2008 or Koeze et al, PRL, 2018 finding fundamentally different physics of jamming with or without adhesion. In the two cited papers the authors only consider equilibrium transitions in systems with attraction using computer simulations. Apparently, jamming properties depend on the strength of attraction. There are no attempts to characterize the dynamics, the focus of our work.

What we meant is that any universal relations, such as the Vogel-FulcherTammann relation, would still be valid. Of course, non-universal quantities such as glass transition temperature Tg or fragility will change. In our case, changing the adhesion strength would change ϕS, and the parameters in the VFT. However, our contention is that the overall finding that increase in viscosity followed by saturation is unlikely to change. We have added some clarifying statements in the manuscript to make this clear.

Version 3

Abstract

A recent experiment on zebrafish blastoderm morphogenesis showed that the viscosity (η) of a non-confluent embryonic tissue grows sharply until a critical cell packing fraction (ϕS). The increase in η up to ϕS is similar to the behavior observed in several glass forming materials, which suggests that the cell dynamics is sluggish or glass-like. Surprisingly, η is a constant above ϕS. To determine the mechanism of this unusual dependence of η on ϕ, we performed extensive simulations using an agent-based model of a dense non-confluent two-dimensional tissue. We show that polydispersity in the cell size, and the propensity of the cells to deform, results in the saturation of the available free area per cell beyond a critical packing fraction. Saturation in the free space not only explains the viscosity plateau above ϕS but also provides a relationship between equilibrium geometrical packing to the dramatic increase in the relaxation dynamics.

Introduction

There is great interest in characterizing the mechanical and dynamical properties of embryonic tissues because they regulate embryo development [1–5]. Measurements of bulk properties, such as viscosity and elastic modulus, and the dynamics of individual cells through imaging techniques, have been interpreted by adapting concepts developed to describe phase transitions, glass transition, and active matter [6–9]. Several experiments have shown that during embryo morphogenesis, material properties of the tissues change dramatically [10– 14]. Of relevance to our study is a remarkable finding that provided evidence that a phase transition (PT) occurs during zebrafish blastoderm morphogenesis, which was analyzed using rigidity percolation theory [14–17]. The authors also estimated the viscosity (η) of the blastoderm tissue using the micropipette aspiration technique [13, 18]. It was found that change in η is correlated with cell connectivity (⟨C⟩), rising sharply over a narrow range of ⟨C⟩. Surprisingly, a single geometrical quantity, the cell-cell contact topology controls both the rigidity PT as well as changes in η in this non-confluent tissue, thus linking equilibrium and transport properties.

Here, we focus on two pertinent questions that arise from the experiments on zebrafish blastoderm. First, experiments [14] showed that η increases as a function of the cell packing fraction (ϕ) till ϕ ≤ 0.87. The dependence of η on ϕ follows the well-known Vogel-Fulcher-Tamman (VFT) law [19], which predicts that η grows monotonically with ϕ. The VFT law, which is commonly used to analyze the viscosity of a class of glass forming materials [20], is given by η expwhere ϕ0 is a constant. Surprisingly, for packing fractions, ϕ ≥ ϕS ≈ 0.90, η deviates from the VFT law and is independent of ϕ, which cannot be explained using conventional theories for glasses [8, 21]. Second, the experimental data [14] was interpreted using equilibrium rigidity percolation theory [15–17] for an embryonic tissue in which cells undergo random cell divisions. A priori it is unclear why equilibrium concepts should hold in zebrafish morphogenesis, which one would expect is controlled by non-equilibrium processes such as self-propulsion, growth and cell division.

We show that the two conundrums (saturation of η at high packing fractions and the use of equilibrium statistical mechanics in a growing system to explain phase transition) may be rationalized by (i) assuming that the interactions between the cells are soft, (ii) the cell sizes are highly heterogeneous (polydisperse), which is the case in zebrafish blastoderm. Using an agent-based (particle) simulation model of a two-dimensional non-confluent tissue, we explore the consequences of varying ϕ (see section Materials and Methods for the definition) of interacting self-propelled polydisperse soft cells, on η. The central results of our study are: (i) The calculated effective viscosity (for details see Appendix F), for the polydisperse cell system, shows that for ϕ ≤ ϕS ≈ 0.90 the increase in viscosity follows the VFT law. Just as in experiments, η is essentially independent of ϕ at high (≥ ϕS) packing fractions. The unusual dependence of η at ϕ ≥ ϕS is quantitatively explained using the notion of available free area fraction (ϕfree), which is the net void space that can be explored by the cells when they are jammed. At high densities, a given cell requires free space in order to move. The free area is created by movement of the neighboring cells into the available void space. One would intuitively expect that the ϕfree should decrease with increasing packing fractions, due to cell jamming, which should slow down the overall dynamics. Indeed, we find that ϕfree decreases with increasing packing fraction (ϕ) until ϕS. The simulations show that when ϕ exceeds ϕS, the free area ϕfree saturates because the soft cells (characterized by “soft deformable disks”) can overlap with each other, resulting in the collective dynamics of cells becoming independent of ϕ for ϕ ≥ ϕS. As a consequence η saturates at high ϕ. (iii) Cells whose sizes are comparable to the available free area move almost like a particle in a liquid. The motility of small sized cells facilitates adjacent cells to move through multi-cell rearrangement even in a highly jammed environment. The facilitation mechanism, invoked in glassy systems [22] allows large cells to move with low mobility. A cascade of such facilitation processes enable all the cells to remain dynamic even above the onset packing fraction of the phase transition. (iv) We find that the relaxation time does not depend on the waiting time for measurements even in the regime where viscosity saturates. In other words, there is no evidence of aging even in the regime where viscosity saturates. Strikingly, the tissue exhibits ergodic [23] behavior at all densities. The cell-based simulations, which reproduce the salient experimental features, may be described using equilibrium statistical mechanics, thus providing credence to the use of cell contact mechanics to describe both rigidity PT and dynamics in an evolving non-confluent tissue [14].

Results

Experimental results

We first describe the experimental observations, which serve as the basis for carrying out the agent-based simulations. Fig. 1(A) shows the bright-field images of distinct stages during zebrafish morphogenesis. A 2D section of zebrafish blastoderm (Fig. 1 (B)) shows that there is considerable dispersion in cell sizes. The statistical properties of the cell sizes are shown in Fig. 8 (D). Fig. 1 (C) shows that η increases sharply over a narrow ϕ range, and saturates when ϕ exceeds ϕS ≈ 0.90.

Structure and viscosity of non-confluent tissues:

(A) Bright-field single-plane images of an exemplary embryo of zebrafish before (t = −60 min), at the onset (t = 0 min), and after blastoderm spreading (t = 60 min). (B) Snapshot of 2D confocal sections at the 1st–2nd deep-cell layer of the blastoderm at t = 60 min. (A) and (B) are taken from [14]. (C) Viscosity η of zebrafish blastoderm as a function of ϕ in a log-linear scale using the data from [14]. The dashed line is the fit to VFT equation. Note that η does not change significantly beyond ϕ ≥ 0.87. (D) A typical snapshot taken from cell-based simulations for ϕ = 0.93. Cells are colored according to their radii (in μm) (color bar shown on the right). (E) The pair correlation function, g(r), as a function of r for ϕ = 0.93. The vertical dashed line is the position of the first peak (rmax = 17.0 μm). The pair correlation function does not exhibit signs of long-range order. Scale bars in (A) is 100 μm and in (B) is 50 μm. range order. Thus, the polydisperse cell system exhibits liquid-like structure even at the high ϕ.

To account for the results in Fig. 1 (C), we first simulated a mono-disperse system in which all the cells have identical radius (R = 8.5 μm). Because the system crystallizes (Fig. 8 (A) & (B)), we concluded that the dynamics observed in experiments cannot be explained using this model. A 1:1 binary mixture of cells, with different radii gives glass-like behavior for all ϕ, with the relaxation time τα as well as the effective viscosity (defined in Eq. 1) following the VFT behavior (see Appendix B).

Polydispersity and cell-cell interactions

In typical cell tissues, and zebrafish in particular, there is a great dispersion in the cell sizes, which vary in a single tissue by a factor of ∼ 5 − 6 [14] (Fig. 1 (B) and Fig. 8 (D)). In addition, the elastic forces characterizing cell-cell interactions are soft, which implies that the cells can overlap, with rij −(Ri +Rj) < 0 when they are jammed ((Fig. 1 (B) and (Fig. 1 (D)). Thus, both polydispersity as well as soft interactions between the cells must control the relaxation dynamics. To test this proposition, we simulated a highly polydisperse system (PDs) in which the cell sizes vary by a factor of ∼ 8 (Fig. 1 (D) and Fig. 8 (E)).

A simulation snapshot (Fig. 1 (D)) for ϕ = 0.93 shows that different sized cells are well mixed. In other words, the cells do not phase separate. The structure of the tissue can be described using the pair correlation function, , where is the number density, δ is the Dirac delta function, is the position of the ith cell, and the angular bracket ⟨⟩ denotes an average over different ensembles. The g(r) function (Fig. 1 (E)) has a peak around r ∼ 17μm, which is approximately the average diameter of the cells. The absence of peaks in g(r) beyond the second one, suggests there is no long

Effective shear viscosity() as a function of ϕ

A fit of the experimental data for η using the VFT [24, 25] relation in the range ϕ ≤ 0.87 (Fig. 1 (C)) yields ϕ0 ≈ 0.95 and D ≈ 0.51 [19]. The VFT equation for cells, which is related to the Doolittle equation [26] for fluidity that is based on free space available for motion in an amorphous system [27, 28], is η = η exp, where D is the apparent activation energy. In order to compare with experiments, we calculated an effective shear viscosity() for the polydisperse system using a Green-Kubo-type relation [29], The stress tensor Pμν(t) in the above equation is, where μ, ν ∈ (x, y) are the Cartesian components of coordinates, is the force between ith and jth cells and A is the area of the simulation box. Note that should be viewed as a proxy for shear viscosity because it does not contain the kinetic term and the factor is not included in Eqn. (1) because temperature is not a relevant variable in the highly over-damped model for cells considered here.

Plot of as a function of ϕ in Fig. 2 (A), shows qualitatively the same behavior as the estimate of viscosity (using dimensional arguments) made in experiments. Two features about Fig. 1 (C) and Fig. 2 (A) are worth noting. (i) Both simulations and experiments show that up to ϕ ≈ 0.90, follows the VFT relation with ϕ0 ∼ 0.94 and D ∼ 0.5. More importantly is independent of ϕ when ϕ > 0.90. (ii) The values of ϕ0 and D obtained by fitting the experimental estimate of η to the VFT equation and simulation results are almost identical. Moreover, the onset of the plateau packing fraction in simulations and experiments occur at the same value (ϕS ∼ 0.90). The overall agreement with experiments is remarkable given that the model was not created to mimic the zebrafish tissue.

Saturation in viscosity and relaxation time:

(A) Effective viscosity as a function of ϕ, with the solid line being the fit to VFT equation. The inset shows at high ϕ. The dashed line in the inset is the expected behavior assuming that the VFT relation holds at all ϕ. (B) The self-intermediate scattering function Fs(q, t) as a function of t for 0.70 ≤ ϕ ≤ 0.905. The dashed line corresponds to . (C) A similar plot for ϕ > 0.905. (D) The logarithm of the relaxation time τα(s) as a function of ϕ. The VFT fit is given by the dashed line. The inset shows a zoomed-in view for ϕ ≥ ϕS.

To provide additional insights into the dynamics, we calculated the isotropic self-intermediate scattering function, Fs(q, t), Where is the wave vector, and is the position of a cell at time t. The degree of dynamic correlation between two cells can be inferred from the decay of Fs(q, t). The angle bracket ⟨…⟩ is an average over different time origins and different trajectories. We chose where rmax is the position of the first peak in g(r) between all cells (see Fig. 1 (E)). The relaxation time τα is calculated using .

From Figs. 2 (B) & (C), which show Fs(q, t) as a function of t for various ϕ, it is clear that the dynamics become sluggish as ϕ increases. The relaxation profiles exhibit a two step decay with a plateau in the intermediate time scales. The dynamics continues to slow down dramatically until ϕ ≤ 0.90. Surprisingly, the increase in the duration of the plateau in Fs(q, t) ceases when ϕ exceeds ≈ 0.90 (Fig. 2 (C)), a puzzling finding that is also reflected in the dependence of τα on ϕ in Fig. 2 (D). The relaxation time increases dramatically, following the VFT relation, till ϕ ≈ 0.90, and subsequently reaches a plateau (see the inset in Fig. 2 (D)). If the VFT relation continued to hold for all ϕ, as in glasses or in binary mixture of 2D cells (see Appendix B), then the fit yields ϕ0 ≈ 0.95 and D ≈ 0.50. However, the simulations show that τα is nearly a constant when ϕ exceeds 0.90. We should note that the behavior in Fig. 2 (D) differs from the dependence of τα on ϕ for 2D monodisperse polymer rings, used as a model for soft colloids. Simulations [30] showed τα increases till a critical ϕS but it decreases substantially beyond ϕS with no saturation.

Relaxation dynamics of individual cells

Plot of τα as a function of the radius of cells Ri (Fig. 3 (A)) shows nearly eight orders of magnitude change. The size dependence of τα on ϕ is striking. That τα should increase for large sized cells (see the data beyond the vertical dashed line in Fig. 3 (A)) is not unexpected. However, even when cell sizes increase beyond Ri = 4.25 μm, the dispersion in τα is substantial, especially when ϕ exceeds ϕS. The relaxation times for cells with Ri < 4.25 μm are relatively short even though the system as a whole is jammed. For ϕ ≥ 0.90, τα for small sized cells have a weak dependence on ϕ. Although τα for cells with radius < 4 μm is short, it is clear that for a given ϕ (for example ϕ = 0.93) the variations in τα is substantial. In contrast, τα’s for larger cells (R ≥ 7μm) are substantially large, possibly exceeding the typical cell division time in experiments. In what follows, we interpret these results in terms of available free area ⟨Afree⟩ for cells. The smaller sized cells have the largest (RS is the radius of the small cell).

Spectrum of relaxation times:

(A) Scatter plot of relaxation times τα(s) as a function of cell radius. From top to bottom, the plot corresponds to decreasing ϕ. The vertical dashed line is for Ri = 4.25 μm, beyond which the τα changes sharply at high packing fractions. (B) Histogram P (ln(τα)) as a function of ln(τα). Beyond ϕ = 0.90 (ϕS), the histogram peaks do not shift substantially towards a high τα values. (C) For ϕ ≤ ϕS P (ln(τα)) (scaled by P max(ln(τα))) falls on a master curve, as described in the main text. (D) Same as (C) except the results are for ϕ > 0.90. The data deviates from the Gaussian fit, shown by the dashed line.

The effect of jamming on the dramatic increase in τα occurs near Ri ≈ 4.5 μm, which is comparable to the length scale of short range interactions. For ϕ ≤ 0.90, τα increases as the cell size increases. However, at higher packing fractions, even cells of similar sizes show substantial variations in τα, which change by almost 3 − 4 orders of magnitude (see the data around the vertical dashed line for ϕ ≥ 0.915 in Fig. 3 (A)). This is a consequence of large variations in the local density (Fig. 14). Some of the similar-sized cells are trapped in the jammed environment, whereas others are in less crowded regions (see Fig. 14). The spread in τα increases dramatically for ϕ > ϕS (≈ 0.90) and effectively overlap with each other. This is vividly illustrated in the the histogram, P(log(τα)) shown in Fig. 3 (B). For ϕ < ϕs, the peak in P(log(τα)) monotonically shifts to higher log(τα) values. In contrast, when ϕ exceeds ϕS there is overlap in P(log(τα)), which is reflected in the saturation of and τα. There are cells (typically with small sizes) that move faster even in a highly jammed environment (see Fig. 12 (C) and Fig. 13). The motions of the fast-moving cells change the local environment, which effectively facilitates the bigger cells to move in a crowded environment (see Fig. 12 (D) and Fig. 13 and and Movie 1 (ϕ = 0.92 > ϕS) and Movie 2 (ϕ = 0.90 = ϕS)). In contrast, for ϕ = 0.85 < ϕS small and large sized cells move without hindrance because of adequate availability of free area (Movie 3). The videos vividly illustrate the large scale facilitated rearrangements that enable the large sized cells to move.

The dependence of τα on ϕ for ϕ ≤ ϕS (Fig. 2 (D)) implies that the polydisperse cell systems behave as a soft glass in this regime. On theoretical grounds, it was predicted that in glass-forming systems [8]. Remarkably, we found that this prediction is valid in the polydisperse cell system (Fig. 3 (C)). However, above ϕS the predicted relation is not satisfied (see Fig. 3 (D)).

Available free area explains viscosity saturation at high ϕ

We explain the saturation in the viscosity by calculating the available free area per cell, as ϕ increases. In a hard disk system, one would expect that the free area would decrease monotonically with ϕ until it is fully jammed at the close packing fraction (∼ 0.84 [31, 32]). Because the cells are modeled as soft deformable disks, they could overlap with each other even when fully jammed. Therefore, the region where cells overlap create free area in the immediate neighborhood. The extent of overlap (hij) is reflected in distribution P(hij). The width in P(hij) increases with ϕ, and the peak shifts to higher values of hij (Fig. 4 (A)). The mean, ⟨hij⟩ increases with ϕ (Fig. 4 (B)). Thus, even if the cells are highly jammed at ϕ ≈ ϕS, free area is available because of an increase in the overlap between cells (see Fig. 5). When ϕ exceeds ϕS, the mobility of small sized cells facilitates the larger cells to move, as is assumed in the free volume theory of polymer glasses [28, 33–35]. As a result of the motion of small cells, a void is temporarily created, which allows other (possibly large) cells to move. In addition to the release of space, the cells can also interpenetrate (Figs. 4 (A) & (B)). If hij increases, as is the case when the extent of compression increases (Figs. 4 (A)), the available space for nearby cells would also increase. This effect is expected to occur with high probability at ϕS and beyond, resulting in high overlap between the cells. These arguments suggest that the combined effect of polydispersity and cell-cell overlap creates, via the self-propulsion of cells, additional free area that drives larger cells to move even under jammed conditions.

Density dependent cell-cell overlap :

(A) Probability of overlap (hij) between two cells, P (hij), for various ϕ values. The peak in the distribution function shifts to higher values as ϕ increases. (B) Mean ⟨hij⟩ = dhijP (hij) as a function of ϕ. Inset shows a pictorial illustration of h12 between two cells with radii R1 and R2 at a distance r12.

Changes in free area fraction with ϕ:

(A) Voronoi tessellation of cells for ϕ = 0.93 for a single realization. The orange circles represent actual cell sizes. The blue polygons show the Voronoi cell size. (B) Distribution of Voronoi cell size A as a function of ϕ. (C) Mean Voronoi cell size ⟨A⟩ as a function of ϕ. A zoomed-in view for ϕ > 0.860 is shown in the inset. (D) Distribution of free area P (Afree) for all ϕ. The vertical blue dashed line shows that the maximum in the distribution is at Afree ∼ 50μm2. (E) Free area fraction ϕfree as a function of ϕ. Note that ϕfree saturates beyond ϕ = 0.90. An expanded view of the saturated region is shown in the right panel of (E).

In order to quantify the physical picture given above, we calculated an effective area for each cell by first calculating Voronoi cell area A. A plot for Voronoi tessellation is presented in Fig. 5 (A) for ϕ = 0.93, and the histogram of A is shown in Fig. 5 (B). As ϕ increases, the distribution shifts towards lower Voronoi cell size ⟨A⟩. The mean Voronoi cell size ⟨A⟩ as a function of ϕ in Fig. 5 (C) shows ⟨A⟩ decreases as ϕ is increased. As cells interpenetrate, the Voronoi cell size will be smaller than the actual cell size in many instances (Fig. 5 (A)). To demonstrate this quantitatively, we calculated . The value of Afree could be negative if the overlap between neighbouring cells is substantial; Afree is positive only when the Voronoi cell size is greater than the actual cell size. Positive Afree is an estimate of the available free area. The histograms of Afree for all the packing fractions in Fig. 5 (D) show that the distributions saturate beyond ϕ = 0.90. All the distributions have a substantial region in which Afree is negative. The negative value of Afree increases with increasing ϕ, which implies that the amount of interpenetration between cells increases.

Because of the overlap between the cells, the available free area fraction ϕfree, is higher than the expected free area fraction (1.0 − ϕ) for all ϕ. We define an effective free area fraction ϕfree as, where Np is the number of positive free area in jth snapshots, Nt is the total number of snapshots, Abox is the simulation box area and is the positive free area of ith cell in jth snapshot. The calculated ϕfree, plotted as a function of ϕ in Fig. 5 (E), shows that ϕfree decreases with ϕ until ϕ = 0.90, and then it saturates near a value ϕfree ≈ 0.22 (see the right panel in Fig. 5 (E)). Thus, the saturation in as a function of ϕ is explained by the free area picture, which arises due to combined effect of the size variations and the ability of cells to overlap.

Aging does not explain viscosity saturation

Our main result, which we explain by adopting the free volume theory developed in the context of glasses [28, 33–35], is that above a critical packing fraction ϕS ∼ 0.90 the viscosity saturates. Relaxation time, τα, measured using dynamic light scattering, in nearly monodisperse microgel poly(N-isopropylacrylamide) (PNiPAM) [36] was found to depend only weakly on the volume fraction (3D), if ϕV, exceeds a critical value. It was suggested that the near saturation of τα at high ϕV is due to aging, which is a non-equilibrium effect. If saturation in viscosity and relaxation time in the embryonic tissue at high ϕ is due to aging then τα should increase sharply as the waiting time, τω, is lengthened. We wondered if aging could explain the observed saturation of η in the embryonic tissue above ϕS. If aging causes the plateau in the tissue dynamics, then η or τα should be an increasing function of the waiting time, τω. To test the effect of τω on τα, we calculated the self-intermediate scattering function Fs(q, t+τω) as a function of t by varying τω over three orders of magnitude at ϕ = 0.92 (Fig. 6 (A)). There is literally no change in Fs(q, t + τω) over the entire range of tω. We conclude that, τα, extracted from Fs(q, t+τω) is independent of τω. The variations in τα (Fig. 6 (B)), with respect to τω, is significantly smaller than the errors in the simulation. Thus, the saturation in η or τα when ϕ > ϕS is not a consequence of aging.

Relaxation in the polydisperse cell system is independent of the waiting time:

(A) Fs(q, t) for ϕ = 0.92 at different waiting times (τω = 106(s)). Regardless of the value of τω, all the Fs(q, t) curves collapse onto a master curve. (B) Relaxation time, ln(τα), as a function of τω. Over a 3 orders of magnitude change in tω, the variation in relaxation times is less than the sample-to-sample fluctuations, as shown by the error bar.

There are two implications related to the absence of aging in the dynamics of the non-confluent embryonic tissues. (i) Although active forces drive the dynamics of the cells, as they presumably do in reality, the cell collectives can be treated as being near equilibrium, justifying the use of Green-Kubo relation to calculate η. (ii) Parenthetically, we note that the absence of significant non-equilibrium effects, even though Zebrafish is a living system, further justifies the use of equilibrium rigidity percolation theory to analyze the experimental data [14].

Discussion

Extensive computer simulations of a two-dimensional dense tissue using a particle-based model of soft deformable cells with active self-propulsion have successfully reproduced the dynamical behavior observed in the blastoderm tissue of zebrafish. The dependence of viscosity (η) and relaxation time (τα) (before the saturation) is well fit by the VFT equation. The value of ϕ0 obtained from simulations, ϕ0 ∼ 0.95, is close to ϕ0 ∼ 0.94 extracted by fitting the experimental data to the VFT equation. Thus, the dynamics for ϕ ≤ ϕS resembles the behavior expected for glass forming systems. Remarkably, the dependence of η on ϕ over the entire range (VFT regime followed by a plateau) may be understood using available free area picture with essentially a single parameter, an idea that was proposed nearly 70 years ago. We discovered that polydispersity as well as the ease of deformation of the cells that creates free area under high jamming conditions, is the mechanism that explains viscosity saturation at high cell densities. The mechanism suggested here is an important step that links equilibrium phase transition to dynamics during zebrafish development [37].

One could legitimately wonder if the extent of polydispersity (PD) used in the soft discs simulations, which seems substantial, is needed to recapitulate the observed dependence of η on ϕ. Furthermore, such large values of PD may not represent biological tissues. Although the choice of PD was made in part by the two-dimensional projection of area reported in experiments [14], it is expected that PD values have to be less in three dimensions. We performed preliminary simulations in three dimensions with considerably reduced PD, and calculated the dependence of relaxation time (τα) as a function of ϕ. The results show that τα does indeed saturate at high volume fractions.

The proposed model neglects adhesive interactions between cells, which of course is not unimportant. It is crucial to wonder if the proposed mechanism would change if cell-cell adhesion is taken into account. We wanted to create the simplest model to explain the ex-perimental data. We do think that realistic values of adhesion strength does not significantly alter the forces between cells [38]. Thus, we expect a similar mechanism. Furthermore, the physics of the dynamics in glass forming materials does not change in systems with and without attractive forces [8]. Universal behavior, such as Vogel-Fucher-Tammann relation, is valid for a broad class of unrelated materials (see Fig.1 in [20]). Needless to say, non-universal quantities such as glass transition temperature Tg or effective free energy barriers for relaxation will change. In our case, we expect that changing the adhesion strength, within a reasonable range, would change ϕS without qualitatively altering the dependence of η on ϕ. For these reasons, in the first pass we neglected adhesion, whose effects have to be investigated in the future.

In the physical considerations leading to Eq. 6, the random activity term (μ) plays an important role. Is it possible to create a passive model by maintaining the system at a finite temperature using stochastic noise with μ = 0, which would show the observed viscosity behavior? First, in such a system of stochastic equations, the coefficient of noise (a diffusion constant) would be related to γi in Eq.6 through fluctuation dissipation theorem (FDT). Thus, only γi can be varied. In contrast, in Eq.6 the two parameters (γi and μ) maybe independently changed, which implies that the two sets of stochastic equations of motion are not equivalent. Second, the passive system describes particles that interact by soft Hertz potential. In analogy with systems in which the particles interact with harmonic potential [39], we expect that the passive model would form a glass in which the viscosity would follow the VFT law.

We find it surprising that the calculation of viscosity using linear response theory (valid for systems close to equilibrium), and the link to free area quantitatively explain the simulation results and by implication the experimental data for a living and growing tissue. The calculation of free area of the cells is based on the geometrical effects of packing, which in turn is determined by cell to cell contact topology. These considerations, that are firmly established here, explain why equilibrium phase transitions are related to steep increase in viscosity [8] as the packing fraction changes over a narrow range. The absence of aging suggests that, although a large number of cell divisions occur, they must be essentially independent, thus allowing the cells to reach local equilibrium.